### 2023-03-22 S1_replicate_stats_and_plots.Rmd
basic analyses ready

---
title: 'Suplementary Material 1: Code to Reproduce Analysis and Plots'
author: "Richard Meitern"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 2)
options(scipen = 9)
options(knitr.kable.NA="")
```

## Introduction

The aim of this document is to display the R code needed the reproduce the findings of the main text. This documents needs the twinR package to be installed to run.

```{r}
#cleanup memory
gcstuff <- gc(verbose=FALSE); rm(gcstuff);
```

```{r}
#get last birth adding function
source("./R/last_birth.R")

#simplified twinR summary tables 
source("./R/twinR_summary.R")

#fix twinR compute predictions to do prediction with no lambda  as well
source("./R/twinR_predictions.R")

#simple convenience functions
source("./R/utils.R")
```

```{r}
## Identify number of CPU cores available for parallel computing,
## note: using a large number may lead RAM to max out, so you may have to adjust
## that according to your infrastructure:
nb_cores <- min(c(50L, parallel::detectCores() - 1))

## Set option in spaMM:
spaMM::spaMM.options(nb_cores = nb_cores)
```

## Data import

The Estonian dataset has been formatted to include the same columns as *the data_births_all* dataset from the **twinR** package. The only difference is that the columns *pop* and *monthly* are excluded as these are constant.

```{r}
#Import and preproccess Estonian Data

data_births_monthly_EE <- readRDS("./data/data_births_all_EE.rds")

#the twinR package expects population to be present
data_births_monthly_EE$pop <- "Estonia"

## Expand the birth level data for the fit of statistical models:
data_births_monthly_EE <- twinR::expand_data(data_births_monthly_EE)

data_births_monthly_EE <- add_last_birth(data_births_monthly_EE)

data_births_monthly_EE_not_last <- data_births_monthly_EE[!data_births_monthly_EE$last,]

#make the aggregates
dmm_EE <- twinR::aggregate_data(data_births_monthly_EE)
dmm_EE$prob_twin <- dmm_EE$twin_total / dmm_EE$births_total

dmm_EE_nl <- twinR::aggregate_data(data_births_monthly_EE_not_last)
dmm_EE_nl$prob_twin <- dmm_EE_nl$twin_total / dmm_EE_nl$births_total
```

```{r}
ee_tbl <- twinR::build_data_summary.table(data_births_monthly_EE)
ee_tbl_nl <- twinR::build_data_summary.table(data_births_monthly_EE_not_last)
```

\tiny
```{r}

  eetbls <- rbind(ee_tbl[-1,4:ncol(ee_tbl)-1],ee_tbl_nl[-1,4:ncol(ee_tbl_nl)-1])
eetbls$Subset <- c("All data", "No last birth")
knitr::kable(eetbls,
             caption = paste0("Data summary table of Estonian data",
                              " with and without last born child")) 
```
\normalsize

## Data import Original

```{r}
##Import and pre-proccess twinR package data

## Filter the raw data to only keep data with monthly resolution:
data_births_monthly <- twinR::filter_data(twinR::data_births_all)

## Expand the birth level data for the fit of statistical models:
data_births_monthly <- twinR::expand_data(data_births_monthly)

data_births_monthly <- add_last_birth(data_births_monthly)
data_births_monthly_not_last <- data_births_monthly[!data_births_monthly$last,]

dmm_orig <- twinR::aggregate_data(data_births_monthly)
dmm_orig$prob_twin <- dmm_orig$twin_total /dmm_orig$births_total

dmm_orig_nl <- twinR::aggregate_data(data_births_monthly_not_last)
dmm_orig_nl$prob_twin <- dmm_orig_nl$twin_total /dmm_orig_nl$births_total

```

\tiny
```{r}
knitr::kable(twinR::build_data_summary.table(data_births_monthly_not_last)[-2],
             caption = "Data sumary table without last born child")
```
\normalsize

## Fitting models

```{r}
#' Fit Predictions
#'
#' This function fits a model using the given formula and dataset and 
#' computes predictions. The model is fit using the \code{\link[spaMM]{fitme}}
#' function from the \strong{spaMM} package.
#' 
#' @param dataset A data frame containing the data to be used for fitting the model.
#' @param formula A formula specifying the model to be fit.
#' @param predict Logical value indicating whether to do predictions. Default is TRUE.
#' @param nb_boot Number of bootstrap samples to use when computing predictions.
#'  Default is 1000.
#' @param predictionsDir Directory where precomputed predictions are stored.
#'  Default is "./data/predictions".
#' 
#' @return A list containing the fitted model object and a data frame with
#'  computed predictions.
fitPredictions <- function(dataset, formula,
                           predict=TRUE,
                           nb_boot=1000,
                           predictionsDir = "./data/predictions"){

  if(!dir.exists(predictionsDir)) dir.create(predictionsDir)

  args <- list(formula = stats::as.formula(formula),
             data = dataset, 
             family = stats::binomial(link = "logit"),
             method = "PQL/L")

  fit <-  twinR::fit_model_safely(timeout = Inf, .args = args)

  fitName <- deparse(substitute(dataset))
  #TODO maybe add formula also to the fitName like:
  #fitName <- paste0(form2str(stats::as.formula(formula)),fitName)

  predDataFname <- paste0(predictionsDir,"/",fitName ,"data_fig.rds")
  if(!file.exists(predDataFname) & predict){
    min_births <- min(dataset$births_total)
    max_births <- max(dataset$births_total)
    nd <- data.frame(births_total = min_births:max_births)
    data_fig <- compute_predictions(fit,
                                    newdata = nd,
                                    nb_boot = nb_boot)

    saveRDS(data_fig, predDataFname)

  } else {
    if(file.exists(predDataFname)){
      warning("Pre-computed predictions returned from file:\n", predDataFname,
            "\n If you want to re-run this time intensive step delete the file!")
    data_fig <- readRDS(predDataFname)
    } else {
      data_fig <- list(results = NULL)
    }

  }

  #garbage collecion after spaMM multi-core proccess
  gcstuff <- gc(verbose=FALSE)

  return(list(fit=fit, results=data_fig$results))
}
```

### Full Data

```{r}
## Estonia - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_fit <- fitPredictions(dmm_EE, formula)

knitr::kable(build_fit_summary.table(dmm_EE_fit$fit))
```

```{r}
## Estonia - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
dbm_EE_fit <- fitPredictions(data_births_monthly_EE, formula, predict = F)

knitr::kable(build_fit_summary.table(dbm_EE_fit$fit))
```

```{r}
## TwinR - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_fit <- fitPredictions(dmm_orig, formula)

knitr::kable(build_fit_summary.table(dmm_orig_fit$fit))
```

```{r}
## TwinR - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
dbm_orig_fit <- fitPredictions(data_births_monthly, formula, predict = F)

knitr::kable(build_fit_summary.table(dbm_orig_fit$fit))
```

### No Last Births Data

```{r}
## Estonia
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_nl_fit <- fitPredictions(dmm_EE_nl, formula)

knitr::kable(build_fit_summary.table(dmm_EE_nl_fit$fit))
```

```{r}
## Estonia - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
dbm_EE_nl_fit <- fitPredictions(data_births_monthly_EE_not_last, formula,
                                predict = F)

knitr::kable(build_fit_summary.table(dbm_EE_nl_fit$fit))
```

```{r}
## TwinR
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_nl_fit <- fitPredictions(dmm_orig_nl, formula)

knitr::kable(build_fit_summary.table(dmm_orig_nl_fit$fit))
```

```{r}
## TwinR - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
dbm_orig_nl_fit <- fitPredictions(data_births_monthly_not_last, formula, predict = F)

knitr::kable(build_fit_summary.table(dbm_orig_nl_fit$fit))
```

## Plots

```{r}
library(ggplot2)
#some nice colors
bc <- c("purple", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")

#use new base theme that displays also grid lines 
source("./R/twinR_theme.R")
```

### Fig 1: Estonian vs TwinR Full Data

```{r}
fig2_EE_plot_data <- dmm_EE_fit$results
fig2_orig_plot_data <- dmm_orig_fit$results
```

```{r}

fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data,
              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
    stat_summary(data=dmm_EE[dmm_EE$births_total <17, ],
                 aes(x=births_total, y=prob_twin, color="EE all", fill = "EE all"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_EE_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    color="EE all", fill = "EE all"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data,
             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
    stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                 aes(x=births_total, y=prob_twin,
                     color="orig. rural", fill="orig. rural"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="orig. rural"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18) + 
    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
    ggplot2::coord_cartesian() +
    labs(subtitle = "Model prediction + data mean with SE",
         y="Per-birth twin. prob.",
         x="Maternal total births")
p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")

p2 
```

### Fig 2: Estonian vs Others No Last Birth

```{r}
fig2_EE_plot_data_nl <- dmm_EE_nl_fit$results
fig2_EE_plot_data_nl$births_total <- fig2_EE_plot_data_nl$births_total + 1
fig2_orig_plot_data_nl <- dmm_orig_nl_fit$results
fig2_orig_plot_data_nl$births_total <- fig2_orig_plot_data_nl$births_total + 1
```

```{r}
dmm_EE_nl_plot <- dmm_EE_nl
dmm_orig_nl_plot <- dmm_orig_nl
dmm_EE_nl_plot$births_total <- dmm_EE_nl_plot$births_total + 1
dmm_orig_nl_plot$births_total <- dmm_orig_nl_plot$births_total + 1
```

```{r}
fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data_nl,
              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
    stat_summary(data=dmm_EE_nl_plot,
                 aes(x=births_total, y=prob_twin, color="EE all", fill = "EE all"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_EE_plot_data_nl,
                aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
                    color="EE all", fill = "EE all"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data_nl,
             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
    stat_summary(data=dmm_orig_nl_plot,
                 aes(x=births_total, y=prob_twin,
                     color="orig. rural", fill="orig. rural"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data_nl,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="orig. rural"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18, limits = c(1,NA)) + 
    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
    ggplot2::coord_cartesian(ylim=c(0,0.03)) +
    labs(subtitle = "Without last children, model prediction + data mean with SE",
         y="Per-birth twin. prob.",
         x="Maternal total births")
p3 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")

p3 
```

```{r}
#END
```

### 2023-03-22 S2_draw_publication_plots.Rmd
basic analyses ready

---
title: 'Suplementary Material 2: Drawing Publication Plots'
author: "Richard Meitern"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 2)
options(scipen = 9)
options(knitr.kable.NA="")
```

## Introduction

The aim of this document is to draw publication plots.

```{r}
#cleanup memory
gcstuff <- gc(verbose=FALSE); rm(gcstuff);
```

```{r}
#get last birth adding function
source("./R/last_birth.R")

#simplified twinR summary tables 
source("./R/twinR_summary.R")

#fix twinR compute predictions to do prediction with no lambda  as well
source("./R/twinR_predictions.R")

#simple convenience functions
source("./R/utils.R")
```

## Data import

```{r}
predDir <- ("./data/predictions/")
dmm_EE_fit <- readRDS(paste0(predDir,"dmm_EEdata_fig.rds"))
dmm_orig_fit <- readRDS(paste0(predDir,"dmm_origdata_fig.rds" ))
dmm_EE_nl_fit <- readRDS(paste0(predDir,"dmm_EE_nldata_fig.rds"))
dmm_orig_nl_fit <- readRDS(paste0(predDir,"dmm_orig_nldata_fig.rds"))
```

```{r}
fig2_EE_plot_data <- dmm_EE_fit$results
fig2_orig_plot_data <- dmm_orig_fit$results
```

```{r}
fig2_EE_plot_data_nl <- dmm_EE_nl_fit$results
fig2_EE_plot_data_nl$births_total <- fig2_EE_plot_data_nl$births_total + 1
fig2_orig_plot_data_nl <- dmm_orig_nl_fit$results
fig2_orig_plot_data_nl$births_total <- fig2_orig_plot_data_nl$births_total + 1
```

```{r}
dmm_EE_nl_plot <- dmm_EE_nl
dmm_orig_nl_plot <- dmm_orig_nl
dmm_EE_nl_plot$births_total <- dmm_EE_nl_plot$births_total + 1
dmm_orig_nl_plot$births_total <- dmm_orig_nl_plot$births_total + 1
```

## Plots

```{r}
library(ggplot2)
#some nice colors
bc <- c("purple", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")

#use new base theme that displays also grid lines 
source("./R/twinR_theme.R")
```

### Fig 1: Estonian vs TwinR Full Data

```{r}

fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data,
              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
    geom_ribbon(data=fig2_EE_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    color="EE all", fill = "EE all"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data,
             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
    geom_ribbon(data=fig2_orig_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="orig. rural"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18) + 
    ggplot2::scale_y_continuous(trans = "logit",
                                breaks = seq(0.005, 0.025, by = 0.001)) +
    ggplot2::coord_cartesian() +
    labs(subtitle = "(A) With last chidren",
         y="Per-birth twin. prob.",
         x="Maternal total births")
p2 <- fig2_ext_orig + base_theme(larger=8, legend_pos = c(0.7,0.9)) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")

p2 
```

### Fig 2: Estonian vs Others No Last Birth

```{r}
fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data_nl,
              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
    geom_ribbon(data=fig2_EE_plot_data_nl,
                aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
                    color="EE all", fill = "EE all"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data_nl,
             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
    geom_ribbon(data=fig2_orig_plot_data_nl,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="orig. rural"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18, limits = c(1,NA)) + 
    ggplot2::scale_y_continuous(trans = "logit",
                                breaks = seq(0.005, 0.025, by = 0.001)) +
    labs(subtitle = "(B) Without last children",
         y="Per-birth twin. prob.",
         x="Maternal total births")
p3 <- fig2_ext_orig + base_theme(larger=8, legend_pos = "none") + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")

p3 
```
```{r, fig.height=8, fig.width=6}
p2_mod <- p2+theme(axis.text.x = element_blank(), axis.title.x = element_blank())
p3_mod <- p3
plot_1 <- gridExtra::grid.arrange(p2_mod, p3_mod, nrow=2)
```

```{r}
#END
```

### 2023-03-22 test.Rmd
basic analyses ready

---
title: "Test"
author: "Richard Meitern"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is an R Markdown test document.

```{r cars}
knitr::kable(summary(cars))
```

## Including Plots
```{r}
library(ggplot2)
```

You can also embed plots, for example:

```{r pressure, echo=FALSE}
ggplot(pressure, aes(x=temperature, y=pressure)) + geom_point()
```

```{r}
#END
```

### 2023-03-22 R/last_birth.R

@@ -1,0 +1,29 @@
+#' Add Column for Last Birth
+#' Find the last birth for all mothers and flag it in the column "last" as TRUE
+#' 
+#' @param dbm data in the same format as from \code{\link[twinR]{data_births_all}}
+#'
+#' @return original data with the column \strong{last} added 
+#'
+#' @examples 
+#' bld <- twinR::filter_data(twinR::data_births_all)
+#' bld_with_last <- add_last_birth(bld)
+add_last_birth <- function(dbm){
+  rows <- nrow(dbm)
+  #get the maximum birth year and mother age for all mothers
+  LBY <- aggregate(cbind(birth_year,maternal_age)  ~  maternal_id, dbm, FUN = max)
+  LBY$last <- TRUE
+  dbm <- merge(dbm, LBY, by= c("birth_year","maternal_age", "maternal_id"), all.x=T)
+  
+  #those that were not maximum set as not last
+  dbm$last <- !is.na(dbm$last)
+  
+  #check for possible duplicates for each mother
+  if(max(table(dbm[dbm$last, c("maternal_id")])) > 1){stop("Aggregate problem!")}
+ 
+  if(nrow(dbm)!=rows){stop("Aggregate problem row mismatch!")}
+  
+  #return in expected order and class
+  dbm <- tibble::as_tibble(dbm)
+  dbm[order(dbm$maternal_id, dbm$birth_year),]
+} 

### 2023-03-22 R/twinR_predictions.R

@@ -1,0 +1,162 @@
+#' Compute the predictions and confidence intervals
+#'
+#' Compute predictions from [`fitme`][`spaMM::fitme`] model fit
+#' with 95% confidence intervals. It is
+#' possible to account for the random effects by integrating predictions over the entire
+#' distribution of such effects. In such a case, the predictions produced thus correspond to
+#' marginal predictions.
+#'
+#' We recommend you to look at the raw R code to understand how they work. We commented the code to make this clear.
+#' While you could directly look at the code of these functions while using the package, mind that
+#' the comments will have been stripped away during the installation process. We kept all the code
+#' of each function self contained, instead of writing modular functions to ease the exploration
+#' of the code (everything is in one place). One drawback is that it leads to a code full of
+#' repetitions.
+#'
+#' @name predictions
+#' @param fit a model fitted (directly or indirectly) with [`fitme`][`spaMM::fitme`] or any [`fit_xx`][`fit_models`] function from this package
+#' @param newdata a `data.frame` providing the values for the fixed effect predictors
+#' @param random a `logical` indicating whether the predictions should be integrated
+#'   over random effect(s)
+#' @param nb_boot the number of simulations for the parametric bootstrap used to
+#'   compute the intervals
+#' @param seed an `integer` providing the seed for the random generator
+#' @return a list containing the predictions  and  intervals, as well as the 
+#' estimates from each bootstrap replicate used to compute 
+#' the intervals and the object created by [`boot.ci`][`boot::boot.ci`].
+#'
+compute_predictions <- function (fit, newdata, random = TRUE, nb_boot = 1000L, seed = 123L) 
+{
+  set.seed(seed)
+  if (nb_boot < 1000L) 
+    warnings("The number of bootstrap simulations used is lower than 1000. You should increase 'nb_boot' for reliable results!")
+  if (length(fit$lambda) > 2L) 
+    stop("This function can only handle up to 2 random effect.")
+  
+  ## identify the response variable:
+  response_var_raw <- paste(fit$call[[2]][[2]])
+  if (length(response_var_raw) == 1L) {
+    response_var <- response_var_raw
+  }
+  else if (length(response_var_raw) > 1L) {
+    response_var <- response_var_raw[2] ## for binomial binary
+  }
+  
+  ## capture the data from the fitted model:
+  data <- fit$data
+  
[…]

### 2023-03-22 R/twinR_summary.R

@@ -1,0 +1,90 @@
+#' Create fit summary tables
+#'
+#' This function is used to create the tables providing the details of the fitted models used in the study.
+#'
+#' @param fit the output of one of the function fitting models (see [`fit_models`])
+#'
+#' @return a `tibble` with an additional attribute called "formula"
+#' @export
+#'
+#' @examples
+#' #See ?twinR
+#'
+build_fit_summary.table <- function(fit) {
+
+  ## extract raw model output:
+  utils::capture.output(fit_components <- spaMM::summary.HLfit(fit))
+
+  ## extract model formula:
+  model_formula <- paste(as.character(fit$call$formula)[c(2,1,3)], collapse = " ")
+
+  ## extract and format info on fixed effects:
+  fixed_effects <- tibble::as_tibble(fit_components$beta_table) |>
+                   dplyr::rename("value" = .data$Estimate) |>
+                   dplyr::mutate(name = rownames(fit_components$beta_table), .before = 1)
+  fixed_effects <- dplyr::bind_cols(object = c("fixed effects", rep("", max(c(0, nrow(fit_components$beta_table) - 1)))), fixed_effects)
+
+  
+  ## extract and format info on random effects:
+  random_effects <- data.frame(object=character(0),name=character(0))
+  if(!is.null(fit_components$lambda_table)){
+    random_effects <- tibble::as_tibble(fit_components$lambda_table) |> 
+      dplyr::select(name = "Group") |> 
+      dplyr::mutate(name = ifelse("name" == "maternal_.", "maternal_id", "name"),
+                    name = paste0("variance between ",  "name"))
+    random_effects <- random_effects  |>
+      dplyr::bind_cols(object = c("random effects",  rep("", max(c(0, nrow(random_effects) - 1)))))
+  }
+  
+  if (!is.null(fit$lambda)) {
+    random_effects <- random_effects |> dplyr::mutate(value = fit$lambda)
+  }
+
+  ## extract and format info on model family:
+  tibble::tibble(object = "response family",
+                 name = as.character(fit$family$family),
+                 value = NA) |>
+    dplyr::mutate(name = ifelse(.data$name == "negbin", "negative binomial", .data$name),
+                  name = paste(.data$name, "with", as.character(fit$family$link), "link")) ->  model_family
+
+  if (fit$family$family == "negbin") {
+    model_family |>
[…]

### 2023-03-22 R/twinR_theme.R

@@ -1,0 +1,40 @@
+#' Theme for plots
+#'
+#' This is the custom ggplot theme that is used.
+#' The theme is built around [`theme_bw`][`ggplot2::theme_bw`] and is inspired by
+#' [`theme_twin`][`twinR::theme_twin`].
+#'
+#' @param base_size base font size
+#' @param base_family base font family
+#' @param base_line_size base size for line elements
+#' @param base_rect_size base size for rect elements
+#' @param larger a scalar indicating by how much to increase text size compared to default 
+#' @param legend_pos a numeric vector of length 2 specifying the x and y coordinates of the legend position.
+#' @param legend_dir a character string specifying the direction of the legend ("horizontal" or "vertical").
+#'
+#' @return the theme
+#' @export
+#'
+base_theme <- function (base_size = 11, base_family = "", base_line_size = base_size/22, 
+base_rect_size = base_size/22, larger = 0, legend_pos = c(0.3,0.9), legend_dir="horizontal") 
+{
+  gray <- "#4D4D4D"
+ black <- "#000000"
+  ggplot2::theme_bw(base_size = base_size, base_family = base_family, 
+                    base_line_size = base_line_size, base_rect_size = base_rect_size) + 
+  ggplot2::theme(line = ggplot2::element_line(colour = gray), 
+    rect = ggplot2::element_rect(fill = "white", colour = NA), 
+    text = ggplot2::element_text(colour = black), axis.ticks = ggplot2::element_line(colour = gray), 
+    legend.key = ggplot2::element_rect(colour = NA), 
+    legend.position = legend_pos,
+    legend.direction = legend_dir,
+    panel.grid.minor = element_blank(),
+    panel.border = ggplot2::element_rect(colour = gray), 
+    strip.background = ggplot2::element_rect(fill = "white",  colour = NA),
+    title = ggplot2::element_text(colour = black, size = 7 + larger, family = base_family),
+    axis.text.x = ggplot2::element_text(colour = black, size = 5 + larger, family = base_family),
+    axis.text.y = ggplot2::element_text(colour = black, size = 5 + larger, family = base_family),
+    axis.title.x = ggplot2::element_text(colour = black, size = 7 + larger, family = base_family, margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0)),
[…]

### 2023-03-22 R/utils.R

@@ -1,0 +1,27 @@
+#' Compute R-squared for GLM
+#'
+#' This function computes the R-squared value for a fitted GLM model.
+#' @param m3 A fitted GLM model object.
+#' @param printIt Logical value indicating whether to print the computed
+#'  R-squared value. Default is TRUE.
+#' @return The computed R-squared value.
+glmR2 <- function(m3, printIt=T){
+  r2 <- with(summary(m3), 1 - deviance/null.deviance)
+  if(printIt){
+    cat(paste("R2=",round(r2, 3)))
+  }
+  invisible(r2)
+}
+
+#' Convert Formula to String
+#'
+#' This function converts a formula object into a string representation.
+#' @param formula A formula object to be converted into a string.
+#' @return A string representation of the input formula.
+form2str <- function(formula){
+  fstr <- paste0(as.character(formula)[2:3], collapse = "~")
+  fstr <- gsub(" ", "", fstr)
+  #remove characters not allowed in Windows file names
+  fstr <- gsub("[<>:\"/\\\\|?*]", "_x_", fstr)
+  fstr
+}

### 2023-03-23 R/fit_models.R
move fitting to function

@@ -1,0 +1,114 @@
+#' Fit Predictions
+#'
+#' This function fits a model using the given formula and dataset and 
+#' computes predictions. The model is fit using the \code{\link[spaMM]{fitme}}
+#' function from the \strong{spaMM} package.
+#' 
+#' @param dataset A data frame containing the data to be used for fitting the model.
+#' @param formula A formula specifying the model to be fit.
+#' @param predict Logical value indicating whether to do predictions. Default is TRUE.
+#' @param nb_boot Number of bootstrap samples to use when computing predictions.
+#'  Default is 1000.
+#' @param save Logical value indicating whether to save model fit and predictions
+#'  to file. Useful if running same fits several time  Default is TRUE.
+#' @param saveDir Directory where precomputed predictions and fits are stored.
+#'  Default is "./data/predictions".
+#' 
+#' @return A list containing the fitted model object and a data frame with
+#'  computed predictions.
+#' @seealso \code{\link[twinR]}
+fitPredictions <- function(dataset, formula,
+                           predict=TRUE,
+                           nb_boot=1000,
+                           save=TRUE,
+                           saveDir = "./data/predictions"){
+  
+  if(!dir.exists(saveDir)) dir.create(saveDir)
+  
+  
+  fitName <- deparse(substitute(dataset))
+  # add formula also to the fitName like:
+  fitName <- paste0(form2str(stats::as.formula(formula)),fitName)
+  
+  args <- list(formula = stats::as.formula(formula),
+               data = dataset, 
+               family = stats::binomial(link = "logit"),
+               method = "PQL/L")
+  
+  fitDataFname <- paste0(saveDir,"/",fitName ,"_fit.rds")
+  if(save){
+    if(!file.exists(fitDataFname)){
+      fit <-  twinR::fit_model_safely(timeout = Inf, .args = args)
+      saveRDS(fit, fitDataFname)
+    } else{
+      warning("Pre-computed fit returned from file:\n", fitDataFname,
+              "\n If you want to re-run this step delete the file",
+              " or change the saveDir!\n")
+      fit <- readRDS(fitDataFname)
+    }
+  } else {
+    fit <-  twinR::fit_model_safely(timeout = Inf, .args = args)
+  }
+  
+  #garbage collecion after spaMM multi-core proccess
+  gcstuff <- gc(verbose=FALSE)
+  
+  get_predictions <- function(predDataFname, fit, dataset, args, save){
[…]

### 2023-03-23 R/twinR_summary.R
add columns to summary table

@@ -9,7 +9,7 @@
 #'
 #' @examples
 #' #See ?twinR
-#'
+#' @seealso code{\link[twinR]{build_fit_summary.table}}
 build_fit_summary.table <- function(fit) {
 
   ## extract raw model output:
@@ -88,3 +88,60 @@ build_fit_summary.table <- function(fit) {
 
   all_results
 }
+
+#' Create data summary tables
+#'
+#' This function is used to create the tables providing the details of the data used in the study.
+#'
+#' @param birth_level_data the dataset to summarize
+#'
+#' @return a `tibble`
+#' @export
+#'
+#' @examples
+#' build_data_summary.table(data_births_all)
+#' @seealso \code{\link[twinR]{build_data_summary.table }}
+build_data_summary.table <- function(birth_level_data) {
+  se <- function(x){
+    sd(x, na.rm = T)/sqrt(length(x))
+  }
+  ## add total births and twinning status to the data:
+  birth_level_data |>
+    dplyr::group_by(.data$pop, .data$maternal_id) |>
+    dplyr::mutate(total_births = dplyr::n(),
+                  twinners     = any(.data$twin),
+                  afb = min(.data$maternal_age)/12,
+                  nontwinners  = all(!.data$twin)) |>
+    dplyr::ungroup() -> data
+  
+  ## notes for ASCII encoding of unicodes:
+  # - a umlaut  = \u00e4
+  # - o umlaut  = \u00f6
+  # - o with stroke = \u00f8
+  # - oe        = \u0153
+  # - per mille = \u2030
+  
+  ## build the part of the table that contains information for all populations combined:
+  tibble::tibble("Populations" = paste0(unique(data$pop), collapse = ", "),
+                 "Dataset" = "",
+                 "Maternal birth period" = paste(min(floor(data$maternal_birthyear), na.rm = TRUE), "-", max(floor(data$maternal_birthyear), na.rm = TRUE), sep = ""),
+                 "1/4 and 3/4 quantiles for maternal birth period" = paste(quantile(data$maternal_birthyear, 1/4, na.rm = TRUE),"-", quantile(data$maternal_birthyear, 3/4, na.rm = TRUE), sep = ""),
+                 "Age at first birth (mean-SE) " =  paste(round(base::mean(data$afb, na.rm = TRUE), digits = 2L), "-", signif(se(data$afb),2), sep = ""),
+                 "Mothers" = length(unique(data$maternal_id)),
[…]

### 2023-03-23 R/twinR_theme.R
move fig 4C drawing to function

@@ -14,7 +14,7 @@
 #'
 #' @return the theme
 #' @export
-#'
+#' @seealso \code{\link[twinR]{theme_twin}}
 base_theme <- function (base_size = 11, base_family = "", base_line_size = base_size/22, 
 base_rect_size = base_size/22, larger = 0, legend_pos = c(0.3,0.9), legend_dir="horizontal") 
 {
@@ -32,9 +32,50 @@ base_rect_size = base_size/22, larger = 0, legend_pos = c(0.3,0.9), legend_dir="
     panel.border = ggplot2::element_rect(colour = gray), 
     strip.background = ggplot2::element_rect(fill = "white",  colour = NA),
     title = ggplot2::element_text(colour = black, size = 7 + larger, family = base_family),
+    legend.text = ggplot2::element_text(colour = black, size = 5 + larger, family = base_family),
     axis.text.x = ggplot2::element_text(colour = black, size = 5 + larger, family = base_family),
     axis.text.y = ggplot2::element_text(colour = black, size = 5 + larger, family = base_family),
     axis.title.x = ggplot2::element_text(colour = black, size = 7 + larger, family = base_family, margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0)),
     axis.title.y = ggplot2::element_text(colour = black, size = 7 + larger, family = base_family, margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 10)),
     plot.margin = ggplot2::margin(3, 1, 3, 1))
+}
+
+#' Draw Figure 4C Like Plot
+#'
+#' This function creates a ggplot object that visualizes per-birth twinning
+#'  probability by maternal age and parity like in the figure 4C in twinR package
+#'
+#' @param data A data frame containing the data to be plotted.
+#' @param larger Numeric value specifying the size of the plot elements.
+#' @param y_lims A numeric vector of length 2 specifying the y-axis limits.
+#' @param legend_pos A numeric vector of length 2 specifying the 
+#' x and y coordinates of the legend.
+#' @param legend_dir Character string specifying the direction of the 
+#' legend ("horizontal" or "vertical").
+#'
+#' @return A ggplot object.
+#' @import ggplot2
+#' @seealso \code{\link[twinR]{draw_fig_4C}}
+draw_fig_4C <- function(data,
+                        larger=8,
+                        y_lims = c(0.01, 0.04),
+                        legend_pos = c(0.2,0.8),
[…]

### 2023-03-23 S2_draw_publication_plots.Rmd
fix plot appearence

@@
 #some nice colors
-bc <- c("purple", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")
+bc <- c( "azure3","purple4", "navy", "darkgoldenrod2", "springgreen3", "black")
 
@@
   geom_line(data=fig2_EE_plot_data,
-              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
+              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
     geom_ribbon(data=fig2_EE_plot_data,
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color="EE all", fill = "EE all"),
+                    color="Estonia", fill = "Estonia"),
                 alpha = 0.3) + 
     geom_line(data=fig2_orig_plot_data,
-             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
+             aes(y = estimate,  x=births_total, color="8 European populations"),  size = 1) +
     geom_ribbon(data=fig2_orig_plot_data,
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="orig. rural"),
+                    fill="8 European populations"),
                 alpha = 0.3) +
@@
     ggplot2::coord_cartesian() +
-    labs(subtitle = "(A) With last chidren",
-         y="Per-birth twin. prob.",
+    labs(subtitle = "(A) Complete datasets",
+         y="Per-birth twinning probability",
          x="Maternal total births")
-p2 <- fig2_ext_orig + base_theme(larger=8, legend_pos = c(0.7,0.9)) + scale_color_manual(values=bc) + 
-  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")
+p2 <- fig2_ext_orig + base_theme(larger=5, legend_pos = c(0.5,0.9)) + scale_color_manual(values=bc) + 
+   scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "")
 
@@
   geom_line(data=fig2_EE_plot_data_nl,
-              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
+              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
     geom_ribbon(data=fig2_EE_plot_data_nl,
                 aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
-                    color="EE all", fill = "EE all"),
+                    color="Estonia", fill = "Estonia"),
                 alpha = 0.3) + 
     geom_line(data=fig2_orig_plot_data_nl,
-             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
+             aes(y = estimate,  x=births_total, color="8 European populations"),  size = 1) +
     geom_ribbon(data=fig2_orig_plot_data_nl,
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="orig. rural"),
+                    fill="8 European populations"),
                 alpha = 0.3) +
@@
                                 breaks = seq(0.005, 0.025, by = 0.001)) +
-    labs(subtitle = "(B) Without last children",
-         y="Per-birth twin. prob.",
+    labs(subtitle = "(B) Without last births",
+         y="Per-birth twinning probability",
          x="Maternal total births")
-p3 <- fig2_ext_orig + base_theme(larger=8, legend_pos = "none") + scale_color_manual(values=bc) + 
-  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")
+p3 <- fig2_ext_orig + base_theme(larger=5, legend_pos = "none") + scale_color_manual(values=bc) + 
+  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "")
 
@@
 ```
-```{r, fig.height=8, fig.width=6}
-p2_mod <- p2+theme(axis.text.x = element_blank(), axis.title.x = element_blank())
-p3_mod <- p3
-plot_1 <- gridExtra::grid.arrange(p2_mod, p3_mod, nrow=2)
+```{r, fig.height=11.7, fig.width=8.3}
+p2_mod <- p2
+p3_mod <- p3 +theme( axis.title.y = element_blank())
+plot_1 <- gridExtra::grid.arrange(p2_mod, p3_mod, ncol=2, nrow=2)
+```
+```{r}
+#export A4 plot
+ggplot2::ggsave("./exports/figure_1_mother_level_data_twinnig_prob.pdf",
+                plot_1, height=11.7, width=8.3)
+ggplot2::ggsave("./exports/figure_1_mother_level_data_twinnig_prob.svg",
+                plot_1, height=11.7, width=8.3)
 ```

### 2023-03-23 R/twinR_predictions.R
reference twinR

@@ -24,7 +24,7 @@
 #' @return a list containing the predictions  and  intervals, as well as the 
 #' estimates from each bootstrap replicate used to compute 
 #' the intervals and the object created by [`boot.ci`][`boot::boot.ci`].
-#'
+#' @seealso \code{\link[twinR]{compute_predictions}}
 compute_predictions <- function (fit, newdata, random = TRUE, nb_boot = 1000L, seed = 123L) 
 {
   set.seed(seed)

### 2023-03-23 S1_replicate_stats_and_plots.Rmd
draw more plots move code to functions

@@
 output:
+  word_document: default
   pdf_document: default
-  word_document: default
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+if(!is.null(knitr::opts_knit$get("rmarkdown.pandoc.to"))){
+ if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "docx"){
+ knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+} 
+}
+
 options(digits = 2)
@@
 
-## Data import
+## Data Import Estonia
 
@@
 
-```{r}
-ee_tbl <- twinR::build_data_summary.table(data_births_monthly_EE)
-ee_tbl_nl <- twinR::build_data_summary.table(data_births_monthly_EE_not_last)
-```
-
-\tiny
-```{r}
-
-  eetbls <- rbind(ee_tbl[-1,4:ncol(ee_tbl)-1],ee_tbl_nl[-1,4:ncol(ee_tbl_nl)-1])
-eetbls$Subset <- c("All data", "No last birth")
-knitr::kable(eetbls,
-             caption = paste0("Data summary table of Estonian data",
-                              " with and without last born child")) 
-```
-\normalsize
-
-## Data import Original
+## Data Import 8 European Populations
 
@@
 
-\tiny
 ```{r}
-knitr::kable(twinR::build_data_summary.table(data_births_monthly_not_last)[-2],
-             caption = "Data sumary table without last born child")
+#compared to twinR original add data on :
+#-age at first birth (AFB)
+#-quantiles range mother birth year
+#-quantiles range offspring birth year
+#-mean + SE total births
+all_tbls <- rbind(build_data_summary.table(data_births_monthly),
+                  build_data_summary.table(data_births_monthly_not_last),
+                  build_data_summary.table(data_births_monthly_EE),
+                  build_data_summary.table(data_births_monthly_EE_not_last))
 ```
-\normalsize
+
+```{r}
+all_tbls$Dataset  <- rep(c("All births", "Without last birth"),2)
+all_tbls$Populations  <- rep(c("8 European", "Estonia"),each=2)
+all_tbls <- as.data.frame(all_tbls)
+row.names(all_tbls) <- paste0(all_tbls$Populations, " populations ",
+                              tolower(all_tbls$Dataset)) 
+knitr::kable(t(all_tbls),
+             caption = paste0("Summary data on the Estonian and other ",
+                               "European populations for both ",
+                              "all births and wihout last birth for same mothers"))
+```
 
@@
 ```{r}
-#' Fit Predictions
-#'
-#' This function fits a model using the given formula and dataset and 
-#' computes predictions. The model is fit using the \code{\link[spaMM]{fitme}}
-#' function from the \strong{spaMM} package.
-#' 
-#' @param dataset A data frame containing the data to be used for fitting the model.
-#' @param formula A formula specifying the model to be fit.
-#' @param predict Logical value indicating whether to do predictions. Default is TRUE.
-#' @param nb_boot Number of bootstrap samples to use when computing predictions.
-#'  Default is 1000.
-#' @param predictionsDir Directory where precomputed predictions are stored.
-#'  Default is "./data/predictions".
-#' 
-#' @return A list containing the fitted model object and a data frame with
-#'  computed predictions.
-fitPredictions <- function(dataset, formula,
-                           predict=TRUE,
-                           nb_boot=1000,
-                           predictionsDir = "./data/predictions"){
-
-  if(!dir.exists(predictionsDir)) dir.create(predictionsDir)
-
-  args <- list(formula = stats::as.formula(formula),
-             data = dataset, 
-             family = stats::binomial(link = "logit"),
-             method = "PQL/L")
-
-  fit <-  twinR::fit_model_safely(timeout = Inf, .args = args)
-
-  fitName <- deparse(substitute(dataset))
-  #TODO maybe add formula also to the fitName like:
-  #fitName <- paste0(form2str(stats::as.formula(formula)),fitName)
-
-  predDataFname <- paste0(predictionsDir,"/",fitName ,"data_fig.rds")
-  if(!file.exists(predDataFname) & predict){
-    min_births <- min(dataset$births_total)
-    max_births <- max(dataset$births_total)
-    nd <- data.frame(births_total = min_births:max_births)
-    data_fig <- compute_predictions(fit,
-                                    newdata = nd,
-                                    nb_boot = nb_boot)
-
-    saveRDS(data_fig, predDataFname)
-
-  } else {
-    if(file.exists(predDataFname)){
-      warning("Pre-computed predictions returned from file:\n", predDataFname,
-            "\n If you want to re-run this time intensive step delete the file!")
-    data_fig <- readRDS(predDataFname)
-    } else {
-      data_fig <- list(results = NULL)
-    }
-
-  }
-
-  #garbage collecion after spaMM multi-core proccess
-  gcstuff <- gc(verbose=FALSE)
-
-  return(list(fit=fit, results=data_fig$results))
-}
+# import the function to do model fit and predictions
+source("./R/fit_models.R")
 ```
@@
 formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
-dmm_EE_fit <- fitPredictions(dmm_EE, formula)
+dmm_EE_fit <- fitPredictions(dmm_EE, formula, predict = T)
 
@@
 formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
-dbm_EE_fit <- fitPredictions(data_births_monthly_EE, formula, predict = F)
+dbm_EE_fit <- fitPredictions(data_births_monthly_EE, formula, nb_boot = 0)
 
@@
 formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
-dmm_orig_fit <- fitPredictions(dmm_orig, formula)
+dmm_orig_fit <- fitPredictions(dmm_orig, formula, predict=T)
 
@@
 formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
-dbm_orig_fit <- fitPredictions(data_births_monthly, formula, predict = F)
+dbm_orig_fit <- fitPredictions(data_births_monthly, formula, nb_boot = 0)
 
@@
 formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
-dmm_EE_nl_fit <- fitPredictions(dmm_EE_nl, formula)
+dmm_EE_nl_fit <- fitPredictions(dmm_EE_nl, formula, predict = T)
 
@@
 dbm_EE_nl_fit <- fitPredictions(data_births_monthly_EE_not_last, formula,
-                                predict = F)
+                                predict = T, nb_boot = 0)
 
@@
 formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
-dmm_orig_nl_fit <- fitPredictions(dmm_orig_nl, formula)
+dmm_orig_nl_fit <- fitPredictions(dmm_orig_nl, formula, predict = T)
 
@@
 formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
-dbm_orig_nl_fit <- fitPredictions(data_births_monthly_not_last, formula, predict = F)
+dbm_orig_nl_fit <- fitPredictions(data_births_monthly_not_last, formula, nb_boot = 0)
 
@@
 #some nice colors
-bc <- c("purple", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")
+bc <- c("azure3","purple4", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")
 
@@
 
-### Fig 1: Estonian vs TwinR Full Data
+### Fig 1A: Estonian vs TwinR Full Data
 
@@
 
-```{r}
-
+```{r, fig.height=11.7/2, fig.width=8.3}
+infoTxt <- "model prediction with 95% CI and data mean SE"
 fig2_ext_orig <- ggplot() +  
   geom_line(data=fig2_EE_plot_data,
-              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
+              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
     stat_summary(data=dmm_EE[dmm_EE$births_total <17, ],
-                 aes(x=births_total, y=prob_twin, color="EE all", fill = "EE all"),
+                 aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
                  alpha=0.5,
@@
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color="EE all", fill = "EE all"),
+                    color="Estonia", fill = "Estonia"),
                 alpha = 0.3) + 
     geom_line(data=fig2_orig_plot_data,
-             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
+             aes(y = estimate,  x=births_total, color="8 European"),  size = 1) +
     stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                  aes(x=births_total, y=prob_twin,
-                     color="orig. rural", fill="orig. rural"),
+                     color="8 European", fill="8 European"),
                  alpha=0.5,
@@
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="orig. rural"),
+                    fill="8 European"),
                 alpha = 0.3) +
@@
     ggplot2::coord_cartesian() +
-    labs(subtitle = "Model prediction + data mean with SE",
+    labs(subtitle = paste0("All births, ", infoTxt),
          y="Per-birth twin. prob.",
@@
 p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
-  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")
+  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "population")
 
@@
 
-### Fig 2: Estonian vs Others No Last Birth
+### Fig 1B: Estonian vs Others No Last Birth
 
@@
 
-```{r}
+```{r, fig.height=11.7/2, fig.width=8.3}
 fig2_ext_orig <- ggplot() +  
   geom_line(data=fig2_EE_plot_data_nl,
-              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
+              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
     stat_summary(data=dmm_EE_nl_plot,
-                 aes(x=births_total, y=prob_twin, color="EE all", fill = "EE all"),
+                 aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
                  alpha=0.5,
@@
                 aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
-                    color="EE all", fill = "EE all"),
+                    color="Estonia", fill = "Estonia"),
                 alpha = 0.3) + 
     geom_line(data=fig2_orig_plot_data_nl,
-             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
+             aes(y = estimate,  x=births_total, color="8 European"),  size = 1) +
     stat_summary(data=dmm_orig_nl_plot,
                  aes(x=births_total, y=prob_twin,
-                     color="orig. rural", fill="orig. rural"),
+                     color="8 European", fill="8 European"),
                  alpha=0.5,
@@
     ggplot2::coord_cartesian(ylim=c(0,0.03)) +
-    labs(subtitle = "Without last children, model prediction + data mean with SE",
+    labs(subtitle = paste0("Without last birth, ", infoTxt),
          y="Per-birth twin. prob.",
@@
 p3 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
-  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")
+  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "population")
 
 p3 
+```
+
+### Figure 2: Model Mean Predictions from Birth Level Data
+
+```{r}
+#use new draw_fig_4C function 
+source("./R/twinR_theme.R")
+```
+
+```{r}
+#extract predictions
+fig4C_EE_plot_data <- dbm_EE_fit$results
+fig4C_EE_plot_data$twin <- fig4C_EE_plot_data$estimates
+
+fig4C_orig_plot_data <- dbm_orig_fit$results
+fig4C_orig_plot_data$twin <- fig4C_orig_plot_data$estimates
+
+fig4C_EE_nl_plot_data <- dbm_EE_nl_fit$results
+fig4C_EE_nl_plot_data$twin <- fig4C_EE_nl_plot_data$estimates
+fig4C_EE_nl_plot_data$parity <- fig4C_EE_nl_plot_data$parity + 1
+
+fig4C_orig_nl_plot_data <- dbm_orig_nl_fit$results
+fig4C_orig_nl_plot_data$twin <- fig4C_orig_nl_plot_data$estimates
+fig4C_orig_nl_plot_data$parity <- fig4C_orig_nl_plot_data$parity + 1
+```
+
+```{r, fig.height=11.7, fig.width=8.3}
+
+rescale_cd <- scale_y_continuous(limits = c(NA, 0.035))
+
+fig2a <- draw_fig_4C(fig4C_orig_plot_data) + 
+  labs(subtitle = "(A) 8 European pop., all births")
+
+fig2b <- draw_fig_4C(fig4C_EE_plot_data) + 
+  labs(subtitle = "(B) Estonia, all births")
+
+fig2c <- draw_fig_4C(fig4C_orig_nl_plot_data, y_lims = c(0.01, 0.035)) + 
+  labs(subtitle = "(C) 8 European pop., without last birth")
+
+fig2d <- draw_fig_4C(fig4C_EE_nl_plot_data,y_lims = c(0.01, 0.035)) + 
+  labs(subtitle = "(D) Estonia, without last birth")
+
+gridExtra::grid.arrange(fig2a + guides(), fig2b, fig2c, fig2d, ncol=2) 
 ```

### 2023-03-27 S1_replicate_stats_and_plots.Rmd
change figure labeles

@@
 
-## Data Import 8 European Populations
+## Data Import 9 Western European Populations
 
@@
 all_tbls$Dataset  <- rep(c("All births", "Without last birth"),2)
-all_tbls$Populations  <- rep(c("8 European", "Estonia"),each=2)
+all_tbls$Populations  <- rep(c("9 Western European", "Estonia"),each=2)
 all_tbls <- as.data.frame(all_tbls)
@@
     geom_line(data=fig2_orig_plot_data,
-             aes(y = estimate,  x=births_total, color="8 European"),  size = 1) +
+             aes(y = estimate,  x=births_total, color="9 W-European"),  size = 1) +
     stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                  aes(x=births_total, y=prob_twin,
-                     color="8 European", fill="8 European"),
+                     color="9 W-European", fill="9 W-European"),
                  alpha=0.5,
@@
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="8 European"),
+                    fill="9 W-European"),
                 alpha = 0.3) +
@@
     geom_line(data=fig2_orig_plot_data_nl,
-             aes(y = estimate,  x=births_total, color="8 European"),  size = 1) +
+             aes(y = estimate,  x=births_total, color="9 W-European"),  size = 1) +
     stat_summary(data=dmm_orig_nl_plot,
                  aes(x=births_total, y=prob_twin,
-                     color="8 European", fill="8 European"),
+                     color="9 W-European", fill="9 W-European"),
                  alpha=0.5,
@@
 fig2a <- draw_fig_4C(fig4C_orig_plot_data) + 
-  labs(subtitle = "(A) 8 European pop., all births")
+  labs(subtitle = "(A) 9 W-European pop., all births")
 
@@
 fig2c <- draw_fig_4C(fig4C_orig_nl_plot_data, y_lims = c(0.01, 0.035)) + 
-  labs(subtitle = "(C) 8 European pop., without last birth")
+  labs(subtitle = "(C) 9 W-European pop., without last birth")
 

### 2023-03-27 S2_draw_publication_plots.Rmd
change figure labeles

@@
 predDir <- ("./data/predictions/")
-dmm_EE_fit <- readRDS(paste0(predDir,"dmm_EEdata_fig.rds"))
-dmm_orig_fit <- readRDS(paste0(predDir,"dmm_origdata_fig.rds" ))
-dmm_EE_nl_fit <- readRDS(paste0(predDir,"dmm_EE_nldata_fig.rds"))
-dmm_orig_nl_fit <- readRDS(paste0(predDir,"dmm_orig_nldata_fig.rds"))
+eeForm <- "cbind(twin_total,singleton_total)~1+births_total" 
+origForm <- "cbind(twin_total,singleton_total)~1+births_total+(1_x_pop)"
+dmm_EE_fit <- readRDS(paste0(predDir, eeForm, "dmm_EEdata_fig.rds"))
+dmm_orig_fit <- readRDS(paste0(predDir, origForm, "dmm_origdata_fig.rds" ))
+dmm_EE_nl_fit <- readRDS(paste0(predDir, eeForm, "dmm_EE_nldata_fig.rds"))
+dmm_orig_nl_fit <- readRDS(paste0(predDir, origForm, "dmm_orig_nldata_fig.rds"))
 ```
@@
 fig2_orig_plot_data_nl$births_total <- fig2_orig_plot_data_nl$births_total + 1
-```
-
-```{r}
-dmm_EE_nl_plot <- dmm_EE_nl
-dmm_orig_nl_plot <- dmm_orig_nl
-dmm_EE_nl_plot$births_total <- dmm_EE_nl_plot$births_total + 1
-dmm_orig_nl_plot$births_total <- dmm_orig_nl_plot$births_total + 1
 ```
@@
   geom_line(data=fig2_orig_plot_data,
-             aes(y = estimate,  x=births_total, color="8 European populations"),  size = 1) +
+            aes(y = estimate,  x=births_total, color="9 W-European populations"),  size = 1) +
   geom_ribbon(data=fig2_orig_plot_data,
               aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="8 European populations"),
+                  fill="9 W-European populations"),
               alpha = 0.3) +
@@
   geom_line(data=fig2_orig_plot_data_nl,
-             aes(y = estimate,  x=births_total, color="8 European populations"),  size = 1) +
+            aes(y = estimate,  x=births_total, color="9 W-European populations"),  size = 1) +
   geom_ribbon(data=fig2_orig_plot_data_nl,
               aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="8 European populations"),
+                  fill="9 W-European populations"),
               alpha = 0.3) +

### 2023-03-29 R/twinR_summary.R
format table digits

@@ -10,7 +10,7 @@
 #' @examples
 #' #See ?twinR
 #' @seealso code{\link[twinR]{build_fit_summary.table}}
-build_fit_summary.table <- function(fit) {
+build_fit_summary.table <- function(fit, signif_digits=3) {
 
   ## extract raw model output:
   utils::capture.output(fit_components <- spaMM::summary.HLfit(fit))
@@ -85,7 +85,18 @@ build_fit_summary.table <- function(fit) {
 
   ## add formula as an atribute of the tibble:
   attr(all_results, "formula") <- model_formula
-
+  #format to signficant digits
+  all_results <- apply(all_results, 2, function(col, s){
+    numcol <- suppressWarnings(as.numeric(col))
+    if(all(is.na(numcol))){return(col)}
+    sapply(numcol, function(x,s){
+      maxval <- 10*s - 1
+      if(is.na(x)){return(x)}
+      if(abs(x)<maxval){return(signif(x, s))}
+      else{round(x)}
+    }, s)
+  }, signif_digits)
+  
   all_results
 }
 

### 2023-03-29 S1_replicate_stats_and_plots.Rmd
change plot colors

@@
 output:
+  pdf_document: default
   word_document: default
-  pdf_document: default
 ---
@@
 
-options(digits = 2)
+options(digits = 9)
 options(scipen = 9)
@@
 #some nice colors
-bc <- c("azure3","purple4", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")
+bc <- c("azure4","purple4", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")
 
@@
 ```{r}
+
 fig2_EE_plot_data <- dmm_EE_fit$results
@@
                  aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
-                 alpha=0.5,
+                 alpha=1,
+                 position = position_nudge(x = -0.1),
                  fun.data=mean_se) +
@@
                      color="9 W-European", fill="9 W-European"),
-                 alpha=0.5,
+                 alpha=1,
+                 shape = 1,
+                 position = position_nudge(x = 0.1),
                  fun.data=mean_se) +
@@
                     fill="9 W-European"),
-                alpha = 0.3) +
+                alpha = 0.1) +
     ggplot2::scale_x_continuous(breaks = 1:18) + 
@@
     ggplot2::coord_cartesian() +
-    labs(subtitle = paste0("All births, ", infoTxt),
+    labs(subtitle = paste0("(A) All births, ", infoTxt),
          y="Per-birth twin. prob.",
@@
                  aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
-                 alpha=0.5,
-                 fun.data=mean_se) +
+                 alpha=1,
+                 fun.data=mean_se,
+                 position = position_nudge(x = -0.1)) +
     geom_ribbon(data=fig2_EE_plot_data_nl,
@@
                      color="9 W-European", fill="9 W-European"),
-                 alpha=0.5,
-                 fun.data=mean_se) +
+                 alpha=1,
+                 shape = 1,
+                 fun.data=mean_se,
+                 position = position_nudge(x = 0.1)) +
     geom_ribbon(data=fig2_orig_plot_data_nl,
@@
                     fill="9 W-European"),
-                alpha = 0.3) +
+                alpha = 0.1) +
     ggplot2::scale_x_continuous(breaks = 1:18, limits = c(1,NA)) + 
@@
     ggplot2::coord_cartesian(ylim=c(0,0.03)) +
-    labs(subtitle = paste0("Without last birth, ", infoTxt),
+    labs(subtitle = paste0("(B) Without last birth, ", infoTxt),
          y="Per-birth twin. prob.",

### 2023-03-31 S1_replicate_stats_and_plots.Rmd
visual adjustments to plots

@@
 
-### Fig 1A: Estonian vs TwinR Full Data
+### Fig 1a: Estonian vs TwinR Full Data
 
@@
     ggplot2::coord_cartesian() +
-    labs(subtitle = paste0("(A) All births, ", infoTxt),
+    labs(subtitle = paste0("(a) All births, ", infoTxt),
          y="Per-birth twin. prob.",
@@
 
-### Fig 1B: Estonian vs Others No Last Birth
+### Fig 1b: Estonian vs Others No Last Birth
 
@@
     ggplot2::coord_cartesian(ylim=c(0,0.03)) +
-    labs(subtitle = paste0("(B) Without last birth, ", infoTxt),
+    labs(subtitle = paste0("(b) Without last birth, ", infoTxt),
          y="Per-birth twin. prob.",
@@
 
-fig2a <- draw_fig_4C(fig4C_orig_plot_data) + 
-  labs(subtitle = "(A) 9 W-European pop., all births")
+fig2b <- draw_fig_4C(fig4C_orig_plot_data) + 
+  labs(subtitle = "(b) 9 W-European, all births")
 
-fig2b <- draw_fig_4C(fig4C_EE_plot_data) + 
-  labs(subtitle = "(B) Estonia, all births")
+fig2a <- draw_fig_4C(fig4C_EE_plot_data) + 
+  labs(subtitle = "(a) Estonia, all births")
 
-fig2c <- draw_fig_4C(fig4C_orig_nl_plot_data, y_lims = c(0.01, 0.035)) + 
-  labs(subtitle = "(C) 9 W-European pop., without last birth")
+fig2d <- draw_fig_4C(fig4C_orig_nl_plot_data, y_lims = c(0.01, 0.035)) + 
+  labs(subtitle = "(d) 9 W-European, without last b.")
 
-fig2d <- draw_fig_4C(fig4C_EE_nl_plot_data,y_lims = c(0.01, 0.035)) + 
-  labs(subtitle = "(D) Estonia, without last birth")
+fig2c <- draw_fig_4C(fig4C_EE_nl_plot_data,y_lims = c(0.01, 0.035)) + 
+  labs(subtitle = "(c) Estonia, without last birth")
 

### 2023-03-31 S2_draw_publication_plots.Rmd
visual adjustments to plots

@@
   ggplot2::coord_cartesian() +
-  labs(subtitle = "(A) Complete datasets",
+  labs(subtitle = "(a) Complete datasets",
        y="Per-birth twinning probability",
@@
                               breaks = seq(0.005, 0.025, by = 0.001)) +
-  labs(subtitle = "(B) Without last births",
+  labs(subtitle = "(b) Without last births",
        y="Per-birth twinning probability",

### 2023-03-31 S1_replicate_stats_and_plots.Rmd
group name adjustments

@@
 
-## Data Import 9 Western European Populations
+## Data Import 9 Other European Populations
 
@@
 ```{r}
+#some labels
+estLab <- "Estonian"
+westLab <- "9 other European"
+```
+
+```{r}
 all_tbls$Dataset  <- rep(c("All births", "Without last birth"),2)
-all_tbls$Populations  <- rep(c("9 Western European", "Estonia"),each=2)
+all_tbls$Populations  <- rep(c(westLab, estLab),each=2)
 all_tbls <- as.data.frame(all_tbls)
@@
 infoTxt <- "model prediction with 95% CI and data mean SE"
+
 fig2_ext_orig <- ggplot() +  
   geom_line(data=fig2_EE_plot_data,
-              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
+              aes(y = estimate,  x=births_total, color=estLab),  size = 1) +
     stat_summary(data=dmm_EE[dmm_EE$births_total <17, ],
-                 aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
+                 aes(x=births_total, y=prob_twin, color=estLab, fill = estLab),
                  alpha=1,
@@
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color="Estonia", fill = "Estonia"),
+                    color=estLab, fill = estLab),
                 alpha = 0.3) + 
     geom_line(data=fig2_orig_plot_data,
-             aes(y = estimate,  x=births_total, color="9 W-European"),  size = 1) +
+             aes(y = estimate,  x=births_total, color=westLab),  size = 1) +
     stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                  aes(x=births_total, y=prob_twin,
-                     color="9 W-European", fill="9 W-European"),
+                     color=westLab, fill=westLab),
                  alpha=1,
@@
                 aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    fill="9 W-European"),
+                    fill=westLab),
                 alpha = 0.1) +
@@
   geom_line(data=fig2_EE_plot_data_nl,
-              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
+              aes(y = estimate,  x=births_total, color=estLab),  size = 1) +
     stat_summary(data=dmm_EE_nl_plot,
-                 aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
+                 aes(x=births_total, y=prob_twin, color=estLab, fill = estLab),
                  alpha=1,
@@
                 aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
-                    color="Estonia", fill = "Estonia"),
+                    color=estLab, fill = estLab),
                 alpha = 0.3) + 
     geom_line(data=fig2_orig_plot_data_nl,
-             aes(y = estimate,  x=births_total, color="9 W-European"),  size = 1) +
+             aes(y = estimate,  x=births_total, color=westLab),  size = 1) +
     stat_summary(data=dmm_orig_nl_plot,
                  aes(x=births_total, y=prob_twin,
-                     color="9 W-European", fill="9 W-European"),
+                     color=westLab, fill=westLab),
                  alpha=1,
@@
 fig2b <- draw_fig_4C(fig4C_orig_plot_data) + 
-  labs(subtitle = "(b) 9 W-European, all births")
+  labs(subtitle = paste0("(b) ",westLab,", all births"))
 
 fig2a <- draw_fig_4C(fig4C_EE_plot_data) + 
-  labs(subtitle = "(a) Estonia, all births")
+  labs(subtitle = paste0("(a) ",estLab,", all births"))
 
 fig2d <- draw_fig_4C(fig4C_orig_nl_plot_data, y_lims = c(0.01, 0.035)) +
-  labs(subtitle = "(d) 9 W-European, without last b.")
+  labs(subtitle = paste0("(d) ",westLab,",  w/o last b."))
 
 fig2c <- draw_fig_4C(fig4C_EE_nl_plot_data,y_lims = c(0.01, 0.035)) + 
-  labs(subtitle = "(c) Estonia, without last birth")
+  labs(subtitle = paste0("(c) ",estLab,",  without last birth"))
 

### 2023-03-31 S2_draw_publication_plots.Rmd
group name adjustments

@@
 ```{r}
+#some labels
+estLab <- "Estonian"
+westLab <- "9 other European"
+```
+
+```{r}
 library(ggplot2)
-#some nice colors
+#some nice colours
 bc <- c( "azure3","purple4", "navy", "darkgoldenrod2", "springgreen3", "black")
@@
   geom_line(data=fig2_EE_plot_data,
-            aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
+            aes(y = estimate,  x=births_total, color=estLab),  size = 1) +
   geom_ribbon(data=fig2_EE_plot_data,
               aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                  color="Estonia", fill = "Estonia"),
+                  color=estLab, fill = estLab),
               alpha = 0.3) + 
   geom_line(data=fig2_orig_plot_data,
-            aes(y = estimate,  x=births_total, color="9 W-European populations"),  size = 1) +
+            aes(y = estimate,  x=births_total, color=westLab),  size = 1) +
   geom_ribbon(data=fig2_orig_plot_data,
               aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                  fill="9 W-European populations"),
+                  fill=westLab),
               alpha = 0.3) +
@@
   geom_line(data=fig2_EE_plot_data_nl,
-            aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
+             aes(y = estimate,  x=births_total, color=estLab),  size = 1) +
   geom_ribbon(data=fig2_EE_plot_data_nl,
               aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
-                  color="Estonia", fill = "Estonia"),
+                  color=estLab, fill = estLab),
               alpha = 0.3) + 
   geom_line(data=fig2_orig_plot_data_nl,
-            aes(y = estimate,  x=births_total, color="9 W-European populations"),  size = 1) +
+            aes(y = estimate,  x=births_total, color=westLab),  size = 1) +
   geom_ribbon(data=fig2_orig_plot_data_nl,
               aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                  fill="9 W-European populations"),
+                  fill=westLab),
               alpha = 0.3) +

### 2023-04-03 S3_run_simulations.Rmd
init simulations

---
title: 'Suplementary Material 3: Run Simulations on the Last Birth Omited Data'
author: "Richard Meitern"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!is.null(knitr::opts_knit$get("rmarkdown.pandoc.to"))){
 if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "docx"){
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
} 
}

options(digits = 9)
options(scipen = 9)
options(knitr.kable.NA="")
```

## Introduction

The aim of this document is to display the R code needed the reproduce the simulations for the PISH scenarios. The letters P, I, S & H are used to indicate which mechanisms were included in each simulation scenario:

  * P indicates that a twinning event affects parity progression.
  * I indicates that a twinning event influences the time between that birth and the next one.
  * S indicates that a mother’s reproductive schedule affects both her likelihood of having twins and her total number of births, creating a link between these two factors.
  * H indicates that there is a connection between twinning propensity and intrinsic fertility due to maternal heterogeneity.

The scenario 0 does not include any of these mechanisms. This documents needs the twinR package to be installed to run.

**NB!** Some of the code e.g reading in the data is same as in S1.

```{r}
#cleanup memory
gcstuff <- gc(verbose=FALSE); rm(gcstuff);
```

```{r}
#get last birth adding function
source("./R/last_birth.R")

#simplified twinR summary tables 
source("./R/twinR_summary.R")

#fix twinR compute predictions to do prediction with no lambda  as well
source("./R/twinR_predictions.R")

#simple convenience functions
source("./R/utils.R")
```

```{r}
## Identify number of CPU cores available for parallel computing,
## note: using a large number may lead RAM to max out, so you may have to adjust
## that according to your infrastructure:
nb_cores <- min(c(20L, parallel::detectCores() - 1))

## Set option in spaMM:
spaMM::spaMM.options(nb_cores = nb_cores)
```

## Data Import Estonia

The Estonian dataset has been formatted to include the same columns as *the data_births_all* dataset from the **twinR** package. The only difference is that the columns *pop* and *monthly* are excluded as these are constant.

```{r}
#Import and preproccess Estonian Data

data_births_monthly_EE <- readRDS("./data/data_births_all_EE.rds")

#the twinR package expects population to be present
data_births_monthly_EE$pop <- "Estonia"

## Expand the birth level data for the fit of statistical models:
data_births_monthly_EE <- twinR::expand_data(data_births_monthly_EE)

data_births_monthly_EE <- add_last_birth(data_births_monthly_EE)

data_births_monthly_EE_not_last <- data_births_monthly_EE[!data_births_monthly_EE$last,]

```

## Data Import 9 Other European Populations

```{r}
##Import and pre-proccess twinR package data

## Filter the raw data to only keep data with monthly resolution:
data_births_monthly <- twinR::filter_data(twinR::data_births_all)

## Expand the birth level data for the fit of statistical models:
data_births_monthly <- twinR::expand_data(data_births_monthly)

data_births_monthly <- add_last_birth(data_births_monthly)
data_births_monthly_not_last <- data_births_monthly[!data_births_monthly$last,]

```

## Running Simulations

```{r}
# import the function to do model fit and predictions
source("./R/fit_models.R")
```

The following is copied directly from twinR documentation and only the saving directories and
input data frames are changed.

```{r}
#------------------------------------------------------------------------------------------------
#-------------------------------- Goodness-of-fit tests -----------------------------------------
#------------------------------------------------------------------------------------------------

### IMPORTANT: while all the steps above should work no matter the operating system and whether
### you are using R within a GUI (e.g. RStudio) or not, the following step is very
### computationally intensive and has been tailored to be used on a Unix system (e.g. Linux) and
### directly within a terminal. It may work in RStudio, but this is not guaranteed, nor
### recommended. If you run this code using Windows, it should fallback to running the
### computation sequentially instead of in parallel across multiple CPU cores, which should work
### fine at the cost of requiring probably weeks of running time.

### Run scenarios one by one and save the output in a rda file:

library(twinR)
library(doSNOW)
nb_cores <- min(c(20L, parallel::detectCores() - 1)) ## only 20 since large memory footprint
timeout <- 24 * 60 * 60 # a Day

baseSlopeDir <- "./exports/slopes_under_scenarios"
baseFitDir <- "./exports/fitted_models"

scenarios_to_do <- c("base_model", "P", "I", "S", "H",
                     "PI", "PS", "PH", "IS", "IH", "SH",
                     "PIS", "PIH", "PSH", "ISH", "PISH")
```

### TwinR data

```{r}
expName <- "_orig_nl"
slopeDir <- paste0(baseSlopeDir, expName)
fitDir <- paste0(baseFitDir, expName)

dir.create(slopeDir)
dir.create(fitDir)

data_births_monthly <- data_births_monthly_not_last

## fit all trios of models in parallel (for linux only, but easy to adjust for other OS):
pbmcapply::pbmclapply(scenarios_to_do, function(scenario) {
  fits <- fit_life_histories(scenario = scenario,
                             birth_level_data = data_births_monthly)
  save(fits, file = paste0(fitDir,"/fits_", scenario, "_obs.rda"), compress = "xz")

  rm(fits)
  }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)

for (scenario in scenarios_to_do) {

   load(file = paste0(fitDir,"/fits_", scenario, "_obs.rda"))

   slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,
                                                    N_replicates_level2 = 49L,
                                                    birth_level_data = data_births_monthly,
                                                    life_history_fits = fits,
                                                    scenario = scenario,
                                                    nb_cores = nb_cores,
                                                    timeout = timeout,
                                                    .log = TRUE, lapply_pkg = "pbmcapply",
                                                    verbose = list(fit = TRUE, simu = FALSE))
   rm(fits)
   name_obj <- paste0("slopes_under_", scenario)
   assign(name_obj, value = slopes_under_scenario)
   save(list = name_obj, file = paste0(slopeDir, name_obj, ".rda"))
   rm(list = name_obj) # remove the object behind the name!
   rm(slopes_under_scenario, scenario, name_obj) # remove the object directly
   gc_stuff <- gc(verbose = FALSE)
}

## Combine simulated slopes:

all_slopes <- combine_simulated_slopes(path_slopes = slopeDir)

## Figure 5:

fig5 <- draw_fig_5(all_slopes, width = 1)
fig5Fname <- paste0("./exports/fig5_", expName, ".")
ggplot2::ggsave(file = paste0(fig5Fname, "pdf"), width = 88, height = 70, units = "mm")
ggplot2::ggsave(file = paste0(fig5Fname, "png"), width = 88, height = 70, units = "mm")

## test scenarios (table S13):

tableS13 <- goodness_of_fit(all_slopes)

write.csv(tableS13, file = paste0(fig5Fname, "csv"))

## computing time (for gof analysis):

computing_time_analysis(all_slopes)

```

### Estonain Data

```{r}
expName <- "_EE_nl"
slopeDir <- paste0(baseSlopeDir, expName)
fitDir <- paste0(baseFitDir, expName)

dir.create(slopeDir)
dir.create(fitDir)

data_births_monthly <- data_births_monthly_EE_not_last

## fit all trios of models in parallel (for linux only, but easy to adjust for other OS):
pbmcapply::pbmclapply(scenarios_to_do, function(scenario) {
  fits <- fit_life_histories(scenario = scenario,
                             birth_level_data = data_births_monthly)
  save(fits, file = paste0(fitDir,"/fits_", scenario, "_obs.rda"), compress = "xz")

  rm(fits)
  }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)

for (scenario in scenarios_to_do) {

   load(file = paste0(fitDir,"/fits_", scenario, "_obs.rda"))

   slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,
                                                    N_replicates_level2 = 49L,
                                                    birth_level_data = data_births_monthly,
                                                    life_history_fits = fits,
                                                    scenario = scenario,
                                                    nb_cores = nb_cores,
                                                    timeout = timeout,
                                                    .log = TRUE, lapply_pkg = "pbmcapply",
                                                    verbose = list(fit = TRUE, simu = FALSE))
   rm(fits)
   name_obj <- paste0("slopes_under_", scenario)
   assign(name_obj, value = slopes_under_scenario)
   save(list = name_obj, file = paste0(slopeDir, name_obj, ".rda"))
   rm(list = name_obj) # remove the object behind the name!
   rm(slopes_under_scenario, scenario, name_obj) # remove the object directly
   gc_stuff <- gc(verbose = FALSE)
}

## Combine simulated slopes:

all_slopes <- combine_simulated_slopes(path_slopes = slopeDir)

## Figure 5:

fig5 <- draw_fig_5(all_slopes, width = 1)
fig5Fname <- paste0("./exports/fig5_", expName, ".")
ggplot2::ggsave(file = paste0(fig5Fname, "pdf"), width = 88, height = 70, units = "mm")
ggplot2::ggsave(file = paste0(fig5Fname, "png"), width = 88, height = 70, units = "mm")

## test scenarios (table S13):

tableS13 <- goodness_of_fit(all_slopes)

write.csv(tableS13, file = paste0(fig5Fname, "csv"))

## computing time (for gof analysis):

computing_time_analysis(all_slopes)

```

```{r}
#END
```

### 2023-04-04 S3_run_simulations_original_not_last.Rmd
refactoring

--- 2023-04-03 S3_run_simulations.Rmd
@@
 gcstuff <- gc(verbose=FALSE); rm(gcstuff);
+
+```
+
+```{r}
+cat("The current directory is: ", getwd())
 ```
@@
 
-```{r}
+```{r make directories}
 expName <- "_orig_nl"
@@
 
-dir.create(slopeDir)
-dir.create(fitDir)
+dir.create(slopeDir, recursive = T)
+dir.create(fitDir, recursive = T)
 
@@
 
+cat("Fits are saved in: ", fitDir)
+```
+
+```{r fit models}
 ## fit all trios of models in parallel (for linux only, but easy to adjust for other OS):
@@
   }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)
+```
 
+```{r run scenarios}
 for (scenario in scenarios_to_do) {
@@
    assign(name_obj, value = slopes_under_scenario)
-   save(list = name_obj, file = paste0(slopeDir, name_obj, ".rda"))
+   save(list = name_obj, file = paste0(slopeDir,"/", name_obj, ".rda"))
    rm(list = name_obj) # remove the object behind the name!
@@
 }
+```
 
+```{r draw figure}
 ## Combine simulated slopes:
@@
 fig5Fname <- paste0("./exports/fig5_", expName, ".")
-ggplot2::ggsave(file = paste0(fig5Fname, "pdf"), width = 88, height = 70, units = "mm")
-ggplot2::ggsave(file = paste0(fig5Fname, "png"), width = 88, height = 70, units = "mm")
+ggplot2::ggsave(filename = paste0(fig5Fname, "pdf"), fig5,
+                width = 88, height = 70, units = "mm")
+fig5
+```
 
+```{r draw figures}
 ## test scenarios (table S13):
@@
 
-write.csv(tableS13, file = paste0(fig5Fname, "csv"))
-
-## computing time (for gof analysis):
-
-computing_time_analysis(all_slopes)
-
+writexl::write_xlsx(tableS13, paste0(fig5Fname, "xlsx"))
+knitr::kable(tableS13)
 ```
 
-### Estonain Data
-
-```{r}
-expName <- "_EE_nl"
-slopeDir <- paste0(baseSlopeDir, expName)
-fitDir <- paste0(baseFitDir, expName)
-
-dir.create(slopeDir)
-dir.create(fitDir)
-
-data_births_monthly <- data_births_monthly_EE_not_last
-
-## fit all trios of models in parallel (for linux only, but easy to adjust for other OS):
-pbmcapply::pbmclapply(scenarios_to_do, function(scenario) {
-  fits <- fit_life_histories(scenario = scenario,
-                             birth_level_data = data_births_monthly)
-  save(fits, file = paste0(fitDir,"/fits_", scenario, "_obs.rda"), compress = "xz")
-
-  rm(fits)
-  }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)
-
-for (scenario in scenarios_to_do) {
-
-   load(file = paste0(fitDir,"/fits_", scenario, "_obs.rda"))
-
-   slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,
-                                                    N_replicates_level2 = 49L,
-                                                    birth_level_data = data_births_monthly,
-                                                    life_history_fits = fits,
-                                                    scenario = scenario,
-                                                    nb_cores = nb_cores,
-                                                    timeout = timeout,
-                                                    .log = TRUE, lapply_pkg = "pbmcapply",
-                                                    verbose = list(fit = TRUE, simu = FALSE))
-   rm(fits)
-   name_obj <- paste0("slopes_under_", scenario)
-   assign(name_obj, value = slopes_under_scenario)
-   save(list = name_obj, file = paste0(slopeDir, name_obj, ".rda"))
-   rm(list = name_obj) # remove the object behind the name!
-   rm(slopes_under_scenario, scenario, name_obj) # remove the object directly
-   gc_stuff <- gc(verbose = FALSE)
-}
-
-## Combine simulated slopes:
-
-all_slopes <- combine_simulated_slopes(path_slopes = slopeDir)
-
-## Figure 5:
-
-fig5 <- draw_fig_5(all_slopes, width = 1)
-fig5Fname <- paste0("./exports/fig5_", expName, ".")
-ggplot2::ggsave(file = paste0(fig5Fname, "pdf"), width = 88, height = 70, units = "mm")
-ggplot2::ggsave(file = paste0(fig5Fname, "png"), width = 88, height = 70, units = "mm")
-
-## test scenarios (table S13):
-
-tableS13 <- goodness_of_fit(all_slopes)
-
-write.csv(tableS13, file = paste0(fig5Fname, "csv"))
-
+```{r timing}
 ## computing time (for gof analysis):

### 2023-04-04 test.Rmd
refactoring

@@
 ```{r cars}
+testDir <- "./exports/test/"
 knitr::kable(summary(cars))
+dir.create(testDir, recursive = T)
+writexl::write_xlsx(cars, "./exports/test/test_cars.xlsx")
 ```
@@
 ## Including Plots
-```{r}
-library(ggplot2)
-```
 
@@
 ```{r pressure, echo=FALSE}
-ggplot(pressure, aes(x=temperature, y=pressure)) + geom_point()
+ggplot2::ggplot(pressure, ggplot2::aes(x=temperature, y=pressure)) + ggplot2::geom_point()
 ```

### 2023-04-04 S3_run_simulations_EE_not_last.Rmd
remove population from default fits

--- 2023-04-04 S3_run_simulations_original_not_last.Rmd
@@
 
-## Expand the birth level data for the fit of statistical models:
-data_births_monthly_EE <- twinR::expand_data(data_births_monthly_EE)
-
 data_births_monthly_EE <- add_last_birth(data_births_monthly_EE)
@@
 data_births_monthly <- twinR::filter_data(twinR::data_births_all)
-
-## Expand the birth level data for the fit of statistical models:
-data_births_monthly <- twinR::expand_data(data_births_monthly)
 
@@
 ```
+```{r}
+#use models without populations predictor
+source("./R/twinR_models.R")
+```
 
@@
 ```{r make directories}
-expName <- "_orig_nl"
+expName <- "_EE_nl"
 slopeDir <- paste0(baseSlopeDir, expName)
@@
 
-data_births_monthly <- data_births_monthly_not_last
+data_births_monthly <- data_births_monthly_EE_not_last
 

### 2023-04-04 S3_run_simulations_EE_not_last.Rmd
run only failed parts

@@
 pbmcapply::pbmclapply(scenarios_to_do, function(scenario) {
+  targetFname <-  paste0(fitDir,"/fits_", scenario, "_obs.rda")
+  if(!file.exists(targetFname)){
     fits <- fit_life_histories(scenario = scenario,
                              birth_level_data = data_births_monthly)
-  save(fits, file = paste0(fitDir,"/fits_", scenario, "_obs.rda"), compress = "xz")
+    save(fits, file =targetFname, compress = "xz")
 
     rm(fits)
+  }
+
   }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)
@@
 
+   name_obj <- paste0("slopes_under_", scenario)
+   targetFname <- paste0(slopeDir,"/", name_obj, ".rda")
+   if(!file.exists(targetFname)){
      slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,
@@
      rm(fits)
-   name_obj <- paste0("slopes_under_", scenario)
      assign(name_obj, value = slopes_under_scenario)
-   save(list = name_obj, file = paste0(slopeDir,"/", name_obj, ".rda"))
+     save(list = name_obj, file = targetFname)
      rm(list = name_obj) # remove the object behind the name!
@@
      gc_stuff <- gc(verbose = FALSE)
+   }
+
 }

### 2023-04-04 S3_run_simulations_original_not_last.Rmd
run only failed parts

@@
 pbmcapply::pbmclapply(scenarios_to_do, function(scenario) {
+  targetFname <-  paste0(fitDir,"/fits_", scenario, "_obs.rda")
+  if(!file.exists(targetFname)){
     fits <- fit_life_histories(scenario = scenario,
                                birth_level_data = data_births_monthly)
-  save(fits, file = paste0(fitDir,"/fits_", scenario, "_obs.rda"), compress = "xz")
+    save(fits, file = targetFname, compress = "xz")
 
     rm(fits)
+  }
   }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)

### 2023-04-05 S3_run_simulations_EE_not_last.Rmd
aproximate RAM usage

@@
 
+The memory (RAM) and usage is quite high for simulations.
+Below there are some information about the RAM  usages per  CPU of simulations on Estonian data w/o last birth.
+The total number of CPUs used for running these simulations was 61.
+
+Memory usage per CPU:
+
+  * base model: about 15GB (dumped cores have a size range 12-23GB)
+
 ```{r}
@@
 ## that according to your infrastructure:
-nb_cores <- min(c(20L, parallel::detectCores() - 1))
+nb_cores <- min(c(200L, parallel::detectCores() - 1))
+
+#it seems that simulating the base model for Estonian not last birth data
+#requires about 15GB of RAM per CPU core (tested with 61 CPU cores)
 
@@
 library(doSNOW)
-nb_cores <- min(c(20L, parallel::detectCores() - 1)) ## only 20 since large memory footprint
+#it seems that simulating the base model for Estonian not last birth data
+#requires about 15GB of RAM per CPU core (tested with 61 CPU cores)
+
 timeout <- 24 * 60 * 60 # a Day
@@
      slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,
-                                                      N_replicates_level2 = 49L,
+                                                      N_replicates_level2 = 20L,
                                                      birth_level_data = data_births_monthly,

### 2023-04-05 S3_run_simulations_EE_not_last.Rmd
remove clutter

@@
 
-## Data Import 9 Other European Populations
-
-```{r}
-##Import and pre-proccess twinR package data
-
-## Filter the raw data to only keep data with monthly resolution:
-data_births_monthly <- twinR::filter_data(twinR::data_births_all)
-
-data_births_monthly <- add_last_birth(data_births_monthly)
-data_births_monthly_not_last <- data_births_monthly[!data_births_monthly$last,]
-
-```
-
 ## Running Simulations
@@
 data_births_monthly <- data_births_monthly_EE_not_last
+
+#remove unnecessary data frames
+rm(data_births_monthly_EE, data_births_monthly_EE_not_last)
 

### 2023-04-05 S3_run_simulations_EE_not_last.Rmd
fix nr of cores

@@
 ## that according to your infrastructure:
-nb_cores <- min(c(200L, parallel::detectCores() - 1))
+nb_cores <- 50
 
@@
 
-timeout <- 24 * 60 * 60 # a Day
+timeout <- 1 * 60 * 60 # an Hour
 

### 2023-04-10 S3_run_simulations_EE_not_last.Rmd
remove unneccesary model loading

@@
 ## that according to your infrastructure:
-nb_cores <- 50
+nb_cores <- 40
 
@@
 for (scenario in scenarios_to_do) {
-
-   load(file = paste0(fitDir,"/fits_", scenario, "_obs.rda"))
-
    name_obj <- paste0("slopes_under_", scenario)
@@
    if(!file.exists(targetFname)){
+     load(file = paste0(fitDir,"/fits_", scenario, "_obs.rda"))
      slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,

### 2023-04-10 S3_run_simulations_EE_not_last.Rmd
make model fitting optional

@@
 ```{r fit models}
+fit_models  <- FALSE
+if(fit_models){
   ## fit all trios of models in parallel (for linux only, but easy to adjust for other OS):
@@
     }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)
+}
+
 ```

### 2023-04-10 S3_run_simulations_EE_not_last.Rmd
updated RAM usage info

@@
 Below there are some information about the RAM  usages per  CPU of simulations on Estonian data w/o last birth.
-The total number of CPUs used for running these simulations was 61.
-
-Memory usage per CPU:
-
-  * base model: about 15GB (dumped cores have a size range 12-23GB)
+The total number of CPUs used for running these simulations was 51.
+RAM usage varied between models but stayed around 18 GB per CPU core (dumped cores had a size range from 12 to 23GB).
 
@@
 
+Its suggested to use a linux computing cluster to run the simulation scenarios one by one (using *run_simulation_EE_not_last.R*) rather than running the for loop below.
+Example of a bash command > **Rscript run_simulation_EE_not_last.R --model H --cores 51**.
+The model indicates the scenario to be run.
+
 ```{r run scenarios}

### 2023-04-10 S3_run_simulations_EE_not_last.Rmd
simulation running explanations

@@
 
-The memory (RAM) and usage is quite high for simulations.
-Below there are some information about the RAM  usages per  CPU of simulations on Estonian data w/o last birth.
-The total number of CPUs used for running these simulations was 51.
+The memory (RAM) usage is quite high for simulations.
+The total number of CPUs used for running single simulation on Estonian data w/o last birth was 51.
 RAM usage varied between models but stayed around 18 GB per CPU core (dumped cores had a size range from 12 to 23GB).

### 2023-04-12 S3_run_simulations_EE_not_last.Rmd
simulations finished

@@
 ## that according to your infrastructure:
-nb_cores <- 40
+nb_cores <- 50
 
@@
 
-```{r draw figure}
+```{r load slopes, include=FALSE}
 ## Combine simulated slopes:
@@
 all_slopes <- combine_simulated_slopes(path_slopes = slopeDir)
+```
 
-## Figure 5:
+```{r}
+#change the p-val expected direction
+source("./R/twinR_goodness_of_fit.R")
 
+# unchanged from the original but needed to have positive slopes yellow. 
+source("./R/twinR_fig_5.R")
+```
+
+```{r draw figure}
+## Figure 5 with colours inverted as we look for positive slopes:
 fig5 <- draw_fig_5(all_slopes, width = 1)
@@
 
-```{r timing}
+```{r timing, results='asis'}
 ## computing time (for gof analysis):
-
-computing_time_analysis(all_slopes)
+timeHours <- computing_time_analysis(all_slopes)
+cat("Total CPU time to run simulations was ",
+    unlist(round(timeHours / 24, 2)), "days. ",
+    "On a high RAM desktop computer with 8 cores it would hence take around",
+    unlist(round(timeHours / 24 / 8, 0)), "days to run. However, on a standard ",
+    "desktop computer with 24 GB RAM only 1 core could be used.")
 

### 2023-04-12 R/twinR_goodness_of_fit.R

@@ -1,0 +1,62 @@
+#' Perform the goodness of fit test of simulation scenario(s)
+#'
+#' This functions perform the goodness of fit test on the data created by the functions
+#' [`simulate_slopes_for_GOF`] or [`combine_simulated_slopes`]. The justification and formal
+#' description of the test are given in the Appendix S1 of our Supplementary Material.
+#'
+#' @param slopes_obj an object returned by [`simulate_slopes_for_GOF`] or [`combine_simulated_slopes`]
+#'
+#' @return a tibble with the information about the scenario and the computed p-values
+#' @export
+#' @seealso [`simulate_slopes_for_GOF`], [`combine_simulated_slopes`]
+#' @examples
+#' # See ?twinR
+#'
+goodness_of_fit <- function(slopes_obj) {
+  
+  ## recursive call if the list contains multiple scenarios:
+  if (length(unique(slopes_obj$scenario)) > 1L) {
+    slopes_obj |>
+      dplyr::mutate(scenario = factor(.data$scenario, levels = c("P", "I", "S", "H", "PI", "PS", "PH", "IS", "IH", "SH", "PIS", "PIH", "PSH", "ISH", "PISH", "base_model"))) |>
+      dplyr::group_split(.data$scenario) -> list_slopes_obj
+    
+    all_gof <- lapply(list_slopes_obj, goodness_of_fit)
+    return(do.call("rbind", all_gof))
+  }
+  
+  ## we do not need the individual slopes for the 2nd level of bootstrapping but only their means,
+  ## so we aggregate the data:
+  slopes_obj |>
+    dplyr::group_by(.data$scenario, .data$seed) |>
+    dplyr::summarize(
+      slope_observed = unique(.data$slope_observed),
+      slopes_level1 = unique(.data$slopes_level1),
+      mean_slopes_level2 = mean(.data$slopes_level2), .groups = "keep") |>
+    dplyr::ungroup() -> slopes_obj_aggregated
+  
+  
+  ## fit a model predicting slopes of first bootstrap based on the slopes at second bootstrap:
+  fit <- stats::lm(slopes_level1 ~ mean_slopes_level2, data = slopes_obj_aggregated)
+  
+  ## compute residuals
+  var_error <- summary(fit)$sigma
+  residuals_slopes_level1 <- var_error*stats::rstudent(fit) # gives similar results than simply using residuals(fit) but more theoretically sound
+  
[…]

### 2023-04-18 S4_draw_IBI_predictions_EE_not_last.Rmd
replicate fig 4

--- 2023-04-10 S3_run_simulations_EE_not_last.Rmd
@@
 ---
-title: 'Suplementary Material 3: Run Simulations on the Last Birth Omited Data'
+title: 'Suplementary Material 4: Draw Interbirth Interval Predictons on the Last Birth Omited Data'
 author: "Richard Meitern"
@@
 
-The aim of this document is to display the R code needed the reproduce the simulations for the PISH scenarios.
-The letters P, I, S & H are used to indicate which mechanisms were included in each simulation scenario:
-
-  * P indicates that a twinning event affects parity progression.
-  * I indicates that a twinning event influences the time between that birth and the next one.
-  * S indicates that a mother’s reproductive schedule affects both her likelihood of having twins and her total number of births, creating a link between these two factors.
-  * H indicates that there is a connection between twinning propensity and intrinsic fertility due to maternal heterogeneity.
-
-The scenario 0 does not include any of these mechanisms. This documents needs the twinR package to be installed to run.
+The aim of this document is to display the R code needed the reproduce the original paper figure 4 equivalent on the Estonian dataset where the last births have been ommited.
 
@@
 
-The memory (RAM) usage is quite high for simulations.
-The total number of CPUs used for running single simulation on Estonian data w/o last birth was 51.
-RAM usage varied between models but stayed around 18 GB per CPU core (dumped cores had a size range from 12 to 23GB).
-
 ```{r}
@@
 ## that according to your infrastructure:
-nb_cores <- 40
+nb_cores <- 7
 
@@
 
-## Running Simulations
+## Get Models
 
 ```{r}
-# import the function to do model fit and predictions
-source("./R/fit_models.R")
+fig_data_files <- sapply(LETTERS[1:3], function(l){
+  predDir <- "./data/predictions/"
+  paste0(predDir, "data_fig_4",l, ".rds")
+})
 ```
 
-The following is copied directly from twinR documentation and only the saving directories and
-input data frames are changed.
+```{r}
+if(!all(sapply(fig_data_files, file.exists))){
+  fitDir <- "D:/hpc/projects/natCom_twinR_commentary/exports/fitted_models_EE_nl/"
+  load(paste0(fitDir, "fits_PISH_obs.rda")) # fit all
 
-```{r}
-#------------------------------------------------------------------------------------------------
-#-------------------------------- Goodness-of-fit tests -----------------------------------------
-#------------------------------------------------------------------------------------------------
+  data_fig_4A <- twinR::prepare_data_fig_4A(fits$fit_PP)
+  gcs <- gc(verbose = F)
+  saveRDS(data_fig_4A, paste0(predDir, "data_fig_4A", ".rds"))
 
-### IMPORTANT: while all the steps above should work no matter the operating system and whether
-### you are using R within a GUI (e.g. RStudio) or not, the following step is very
-### computationally intensive and has been tailored to be used on a Unix system (e.g. Linux) and
-### directly within a terminal. It may work in RStudio, but this is not guaranteed, nor
-### recommended. If you run this code using Windows, it should fallback to running the
-### computation sequentially instead of in parallel across multiple CPU cores, which should work
-### fine at the cost of requiring probably weeks of running time.
+  data_fig_4B <- twinR::prepare_data_fig_4B(fits$fit_IBI)
+  gcs <- gc(verbose = F)
+  saveRDS(data_fig_4B, paste0(predDir, "data_fig_4B", ".rds"))
 
-### Run scenarios one by one and save the output in a rda file:
-
-library(twinR)
-library(doSNOW)
-#it seems that simulating the base model for Estonian not last birth data
-#requires about 15GB of RAM per CPU core (tested with 61 CPU cores)
-
-timeout <- 1 * 60 * 60 # an Hour
-
-baseSlopeDir <- "./exports/slopes_under_scenarios"
-baseFitDir <- "./exports/fitted_models"
-
-scenarios_to_do <- c("base_model", "P", "I", "S", "H",
-                     "PI", "PS", "PH", "IS", "IH", "SH",
-                     "PIS", "PIH", "PSH", "ISH", "PISH")
-```
-```{r}
-#use models without populations predictor
-source("./R/twinR_models.R")
-```
-
-### TwinR data
-
-```{r make directories}
-expName <- "_EE_nl"
-slopeDir <- paste0(baseSlopeDir, expName)
-fitDir <- paste0(baseFitDir, expName)
-
-dir.create(slopeDir, recursive = T)
-dir.create(fitDir, recursive = T)
-
-data_births_monthly <- data_births_monthly_EE_not_last
-
-#remove unnecessary data frames
-rm(data_births_monthly_EE, data_births_monthly_EE_not_last)
-
-cat("Fits are saved in: ", fitDir)
-```
-
-```{r fit models}
-fit_models  <- FALSE
-if(fit_models){
-  ## fit all trios of models in parallel (for linux only, but easy to adjust for other OS):
-  pbmcapply::pbmclapply(scenarios_to_do, function(scenario) {
-    targetFname <-  paste0(fitDir,"/fits_", scenario, "_obs.rda")
-    if(!file.exists(targetFname)){
-      fits <- fit_life_histories(scenario = scenario,
-                               birth_level_data = data_births_monthly)
-      save(fits, file =targetFname, compress = "xz")
-
-      rm(fits)
+  data_fig_4C <- twinR::prepare_data_fig_4C(fits$fit_twinning.binary)
+  gcs <- gc(verbose = F)
+  saveRDS(data_fig_4C, paste0(predDir, "data_fig_4C", ".rds"))
+} else{
+  for(l in LETTERS[1:3]){
+    assign(paste0("data_fig_4",l), readRDS(fig_data_files[l]))
     }
-
-    }, mc.cores = min(c(length(scenarios_to_do), nb_cores)), mc.preschedule = FALSE)
 }
@@
 
-Its suggested to use a linux computing cluster to run the simulation scenarios one by one (using *run_simulation_EE_not_last.R*) rather than running the for loop below.
-Example of a bash command > **Rscript run_simulation_EE_not_last.R --model H --cores 51**.
-The model indicates the scenario to be run.
+## Plot models
 
-```{r run scenarios}
-for (scenario in scenarios_to_do) {
-   name_obj <- paste0("slopes_under_", scenario)
-   targetFname <- paste0(slopeDir,"/", name_obj, ".rda")
-   if(!file.exists(targetFname)){
-     load(file = paste0(fitDir,"/fits_", scenario, "_obs.rda"))
-     slopes_under_scenario <- simulate_slopes_for_GOF(N_replicates_level1 = 200L,
-                                                      N_replicates_level2 = 20L,
-                                                     birth_level_data = data_births_monthly,
-                                                      life_history_fits = fits,
-                                                      scenario = scenario,
-                                                      nb_cores = nb_cores,
-                                                      timeout = timeout,
-                                                      .log = TRUE, lapply_pkg = "pbmcapply",
-                                                      verbose = list(fit = TRUE, simu = FALSE))
-     rm(fits)
-     assign(name_obj, value = slopes_under_scenario)
-     save(list = name_obj, file = targetFname)
-     rm(list = name_obj) # remove the object behind the name!
-     rm(slopes_under_scenario, scenario, name_obj) # remove the object directly
-     gc_stuff <- gc(verbose = FALSE)
-   }
-
-}
+```{r}
+fig_4A <- twinR::draw_fig_4A(data_fig_4A)
+fig_4B <- twinR::draw_fig_4B(data_fig_4B)
+fig_4C <- twinR::draw_fig_4C(data_fig_4C)
 ```
 
-```{r draw figure}
-## Combine simulated slopes:
-
-all_slopes <- combine_simulated_slopes(path_slopes = slopeDir)
-
-## Figure 5:
-
-fig5 <- draw_fig_5(all_slopes, width = 1)
-fig5Fname <- paste0("./exports/fig5_", expName, ".")
-ggplot2::ggsave(filename = paste0(fig5Fname, "pdf"), fig5,
-                width = 88, height = 70, units = "mm")
-fig5
+```{r}
+twinR::plot_grid(fig_4A, fig_4B, fig_4C,
+          labels = "auto", nrow = 1, label_size = 7, align = "v", axis = "lr")
 ```
 
-```{r draw figures}
-## test scenarios (table S13):
-
-tableS13 <- goodness_of_fit(all_slopes)
-
-writexl::write_xlsx(tableS13, paste0(fig5Fname, "xlsx"))
-knitr::kable(tableS13)
-```
-
-```{r timing}
-## computing time (for gof analysis):
-
-computing_time_analysis(all_slopes)
-
+```{r}
+ggplot2::ggsave(file = "./exports/fig4.pdf", width = 180, height = 55, units = "mm")
 ```

### 2023-04-18 S4_draw_IBI_predictions_EE_not_last.Rmd
remve uneccesary stuff explain

@@
 
-The aim of this document is to display the R code needed the reproduce the original paper figure 4 equivalent on the Estonian dataset where the last births have been ommited.
-
-**NB!** Some of the code e.g reading in the data is same as in S1.
-
-```{r}
-#cleanup memory
-gcstuff <- gc(verbose=FALSE); rm(gcstuff);
-
-```
+The aim of this document is to display the R code needed the reproduce the original paper figure 4 equivalent on the Estonian dataset where the last births have been omitted.
 
@@
 cat("The current directory is: ", getwd())
-```
-
-```{r}
-#get last birth adding function
-source("./R/last_birth.R")
-
-#simplified twinR summary tables 
-source("./R/twinR_summary.R")
-
-#fix twinR compute predictions to do prediction with no lambda  as well
-source("./R/twinR_predictions.R")
-
-#simple convenience functions
-source("./R/utils.R")
-```
-
-```{r}
-## Identify number of CPU cores available for parallel computing,
-## note: using a large number may lead RAM to max out, so you may have to adjust
-## that according to your infrastructure:
-nb_cores <- 7
-
-#it seems that simulating the base model for Estonian not last birth data
-#requires about 15GB of RAM per CPU core (tested with 61 CPU cores)
-
-## Set option in spaMM:
-spaMM::spaMM.options(nb_cores = nb_cores)
-```
-
-## Data Import Estonia
-
-The Estonian dataset has been formatted to include the same columns as *the data_births_all* dataset from the **twinR** package.
-The only difference is that the columns *pop* and *monthly* are excluded as these are constant.
-
-```{r}
-#Import and preproccess Estonian Data
-
-data_births_monthly_EE <- readRDS("./data/data_births_all_EE.rds")
-
-#the twinR package expects population to be present
-data_births_monthly_EE$pop <- "Estonia"
-
-data_births_monthly_EE <- add_last_birth(data_births_monthly_EE)
-
-data_births_monthly_EE_not_last <- data_births_monthly_EE[!data_births_monthly_EE$last,]
-
 ```
@@
 ```
+
+As running the *fit_life_histories* function with scenario “**PISH**” will fit all three life history models: *fit_PP*, *fit_IBI*, and *fit_twinning.binary*.
+The polynomial orders for each of these models will be estimated as the one leading to the best fit between 0 and 6.
+The *fit_PP* and *fit_IBI* functions will include the variable twin as a predictor.
+The *fit_twinning.binary* function will include *maternal_id* as a random effect predictor.
+So, it is equivalent to running each of these three functions with their default settings and the specified scenario.
+Hence, to produce the figure 4 equivalent we load these fits and get the fitted models from there.
 
@@
 
+Next we plot the models exactly as in the original manuscript.
+Note that the actual parity is +1 than the number on the graphs (because the last birth is omitted in this data, so the *parity* axis shows the parity up until the one to last).
+
 ```{r}

### 2023-04-18 S4_draw_IBI_predictions_EE_not_last.Rmd
amend explanations

@@
 
-As running the *fit_life_histories* function with scenario “**PISH**” will fit all three life history models: *fit_PP*, *fit_IBI*, and *fit_twinning.binary*.
+Running the *fit_life_histories* function with scenario “**PISH**” will fit all three life history models: *fit_PP*, *fit_IBI*, and *fit_twinning.binary*.
 The polynomial orders for each of these models will be estimated as the one leading to the best fit between 0 and 6.
@@
 So, it is equivalent to running each of these three functions with their default settings and the specified scenario.
-Hence, to produce the figure 4 equivalent we load these fits and get the fitted models from there.
+Hence, to produce the figure 4 equivalent we can load  these fits from the simulation scenario “**PISH**” and get the fitted models *fit_PP*, *fit_IBI*, and *fit_twinning.binary* from there.
 

### 2023-04-18 S4_draw_IBI_predictions_EE_not_last.Rmd
added residuals plot

@@
 ```
+## Data Import Estonia
+
+```{r}
+#get last birth adding function
+source("./R/last_birth.R")
+```
+
+The Estonian dataset has been formatted to include the same columns as *the data_births_all* dataset from the **twinR** package.
+The only difference is that the columns *pop* and *monthly* are excluded as these are constant.
+
+```{r}
+#Import and preproccess Estonian Data
+
+data_births_monthly_EE <- readRDS("./data/data_births_all_EE.rds")
+
+#the twinR package expects population to be present
+data_births_monthly_EE$pop <- "Estonia"
+
+## Expand the birth level data for the fit of statistical models:
+data_births_monthly_EE <- twinR::expand_data(data_births_monthly_EE)
+
+data_births_monthly_EE <- add_last_birth(data_births_monthly_EE)
+
+data_births_monthly_EE_not_last <- data_births_monthly_EE[!data_births_monthly_EE$last,]
+
+#make the aggregates
+dmm_EE <- twinR::aggregate_data(data_births_monthly_EE)
+dmm_EE$prob_twin <- dmm_EE$twin_total / dmm_EE$births_total
+
+dmm_EE_nl <- twinR::aggregate_data(data_births_monthly_EE_not_last)
+dmm_EE_nl$prob_twin <- dmm_EE_nl$twin_total / dmm_EE_nl$births_total
+
+#remove the unused
+rm(data_births_monthly_EE, dmm_EE, data_births_monthly_EE_not_last)
+```
 
@@
 ```{r}
-fig_data_files <- sapply(LETTERS[1:3], function(l){
 predDir <- "./data/predictions/"
-  paste0(predDir, "data_fig_4",l, ".rds")
+fig_data_files <- sapply(LETTERS[1:4], function(l){
+  paste0(predDir, "data_fig_",l, ".rds")
 })
@@
 if(!all(sapply(fig_data_files, file.exists))){
+  #wherever you have stored the fits 
   fitDir <- "D:/hpc/projects/natCom_twinR_commentary/exports/fitted_models_EE_nl/"
-  load(paste0(fitDir, "fits_PISH_obs.rda")) # fit all
 
-  data_fig_4A <- twinR::prepare_data_fig_4A(fits$fit_PP)
+  load(paste0(fitDir, "fits_PISH_obs.rda")) # load all fits all ~ 8 GB
+
+  data_fig_A <- twinR::prepare_data_fig_4A(fits$fit_PP)
   gcs <- gc(verbose = F)
-  saveRDS(data_fig_4A, paste0(predDir, "data_fig_4A", ".rds"))
+  saveRDS(data_fig_A, paste0(predDir, "data_fig_A", ".rds"))
 
-  data_fig_4B <- twinR::prepare_data_fig_4B(fits$fit_IBI)
+  data_fig_B <- twinR::prepare_data_fig_4B(fits$fit_IBI)
   gcs <- gc(verbose = F)
-  saveRDS(data_fig_4B, paste0(predDir, "data_fig_4B", ".rds"))
+  saveRDS(data_fig_B, paste0(predDir, "data_fig_B", ".rds"))
 
-  data_fig_4C <- twinR::prepare_data_fig_4C(fits$fit_twinning.binary)
+  data_fig_C <- twinR::prepare_data_fig_4C(fits$fit_twinning.binary)
   gcs <- gc(verbose = F)
-  saveRDS(data_fig_4C, paste0(predDir, "data_fig_4C", ".rds"))
+  saveRDS(data_fig_C, paste0(predDir, "data_fig_C", ".rds"))
+
+  data_fig_D <- twinR::prepare_data_fig_S2(fit_PP = fits$fit_PP,
+                                           fit_IBI = fits$fit_IBI,
+                                           fit_twin = fits$fit_twinning.binary,
+                                           mother_level_data = dmm_EE_nl)
+  gcs <- gc(verbose = F)
+  saveRDS(data_fig_D, paste0(predDir, "data_fig_D", ".rds"))
+  rm(fits)
 } else{
-  for(l in LETTERS[1:3]){
-    assign(paste0("data_fig_4",l), readRDS(fig_data_files[l]))
+  for(l in LETTERS[1:4]){
+    assign(paste0("data_fig_",l), readRDS(fig_data_files[l]))
     }
@@
 ```{r}
-fig_4A <- twinR::draw_fig_4A(data_fig_4A)
-fig_4B <- twinR::draw_fig_4B(data_fig_4B)
-fig_4C <- twinR::draw_fig_4C(data_fig_4C)
+fig_4A <- twinR::draw_fig_4A(data_fig_A)
+fig_4B <- twinR::draw_fig_4B(data_fig_B)
+fig_4C <- twinR::draw_fig_4C(data_fig_C)
 ```
@@
 ```{r}
+fig_S2 <- twinR::draw_fig_S2(data_fig_D)
+
+ggplot2::ggsave(file = "./exports/figS2.pdf", width = 188, height = 110, units = "mm")
+ggplot2::ggsave(file = "./exports/figS2.png", width = 188, height = 110, units = "mm")
+```
+
+```{r}
 #END

### 2023-04-18 S4_draw_IBI_predictions_EE_not_last.Rmd
added residuals plot

@@
 
-  load(paste0(fitDir, "fits_PISH_obs.rda")) # load all fits all ~ 8 GB
+  #load(paste0(fitDir, "fits_PISH_obs.rda")) # load all fits all ~ 8 GB
 
@@
 
-ggplot2::ggsave(file = "./exports/figS2.pdf", width = 188, height = 110, units = "mm")
-ggplot2::ggsave(file = "./exports/figS2.png", width = 188, height = 110, units = "mm")
+ggsave(file = "./exports/figS2.pdf", width = 188, height = 110, units = "mm")
 ```

### 2023-12-27 S3_response_to_reviewers.Rmd
started response to reviewers

---
title: 'Suplementary Material 3: Code to Respond to Reviewers'
author: "Richard Meitern"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!is.null(knitr::opts_knit$get("rmarkdown.pandoc.to"))){
 if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "docx"){
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
} 
}

options(digits = 9)
options(scipen = 9)
options(knitr.kable.NA="")
```

## Introduction

The aim of this document is to respond to the reviewers comments. The comments are:

1) I would like to see whether indeed mothers are statistically more likely to stop having children after a twin birth (perhaps I missed that analysis and it was indeed presented)?

2) Directly test the odds that a mother terminate reproduction after a twin birth? In other words, use statistics to assess post-twinning reproductive outcomes compared to post-singleton reproductive outcomes.

3) Plot the real data as a cloud of points behind these lines- it is important to see the actual dataset in analyses like these.

4) Look at the analyses when limiting to opposite-sex offspring?

5) Based on analyses of Finnish data from 1953-1964 Eriksson and Fellman show that illegitimate maternities have a higher twinning rate than legitimate ones when maternal age and birth order are taken into account. They speculate that unmarried mothers are women who tend to become pregnant more easily than others owing to the effect of certain factors, e.g. a higher ovulation rate and to have relatively better physical qualifications for going through a twin pregnancy than mothers in general. 
It seems not unlikely that such mothers may remain single of the twin pregnancy, especially in earlier times?

```{r}
#cleanup memory
gcstuff <- gc(verbose=FALSE); rm(gcstuff);
```

```{r}
#get last birth adding function
source("./R/last_birth.R")

#simplified twinR summary tables 
source("./R/twinR_summary.R")

#fix twinR compute predictions to do prediction with no lambda  as well
source("./R/twinR_predictions.R")

#simple convenience functions
source("./R/utils.R")
```

```{r}
## Identify number of CPU cores available for parallel computing,
## note: using a large number may lead RAM to max out, so you may have to adjust
## that according to your infrastructure:
nb_cores <- min(c(50L, parallel::detectCores() - 1))

## Set option in spaMM:
spaMM::spaMM.options(nb_cores = nb_cores)
```

## Data Import

The Estonian dataset has been formatted to include the same columns as *the data_births_all* dataset from the **twinR** package. The only difference is that the columns *pop* and *monthly* are excluded as these are constant.

```{r}
#some labels
estLab <- "Estonian"
westLab <- "9 other European"
```

```{r}
#Import and preproccess Estonian Data

data_births_monthly_EE <- readRDS("./data/data_births_all_EE.rds")

#the twinR package expects population to be present
data_births_monthly_EE$pop <- "Estonia"

## Expand the birth level data for the fit of statistical models:
data_births_monthly_EE <- twinR::expand_data(data_births_monthly_EE)

data_births_monthly_EE <- add_last_birth(data_births_monthly_EE)

data_births_monthly_EE_not_last <- data_births_monthly_EE[!data_births_monthly_EE$last,]

#make the aggregates
dmm_EE <- twinR::aggregate_data(data_births_monthly_EE)
dmm_EE$prob_twin <- dmm_EE$twin_total / dmm_EE$births_total

dmm_EE_nl <- twinR::aggregate_data(data_births_monthly_EE_not_last)
dmm_EE_nl$prob_twin <- dmm_EE_nl$twin_total / dmm_EE_nl$births_total
```

```{r}
##Import and pre-proccess twinR package data

## Filter the raw data to only keep data with monthly resolution:
data_births_monthly <- twinR::filter_data(twinR::data_births_all)

## Expand the birth level data for the fit of statistical models:
data_births_monthly <- twinR::expand_data(data_births_monthly)

data_births_monthly <- add_last_birth(data_births_monthly)
data_births_monthly_not_last <- data_births_monthly[!data_births_monthly$last,]

dmm_orig <- twinR::aggregate_data(data_births_monthly)
dmm_orig$prob_twin <- dmm_orig$twin_total /dmm_orig$births_total

dmm_orig_nl <- twinR::aggregate_data(data_births_monthly_not_last)
dmm_orig_nl$prob_twin <- dmm_orig_nl$twin_total /dmm_orig_nl$births_total

```

```{r}
#compared to twinR original add data on :
#-age at first birth (AFB)
#-quantiles range mother birth year
#-quantiles range offspring birth year
#-mean + SE total births
all_tbls <- rbind(build_data_summary.table(data_births_monthly),
                  build_data_summary.table(data_births_monthly_not_last),
                  build_data_summary.table(data_births_monthly_EE),
                  build_data_summary.table(data_births_monthly_EE_not_last))
```

## Fitting models

### Comment 1: Logistic regression

We can test the hypothesis that mothers are statistically more likely to stop having children after a twin birth by fitting a logistic regression model with the response variable being the binary variable *last* and the predictor variable being the binary variable *twin*.

```{r}
#fit logistic regression Estonia
logit_fit_EE <- glm(last ~ twin,
                 data = data_births_monthly_EE,
                 family = binomial(link = "logit"))
summary(logit_fit_EE)
```

```{r}
#fit logistic regression Other European
logit_fit <- glm(last ~ twin,
                 data = data_births_monthly,
                 family = binomial(link = "logit"))
summary(logit_fit)
```
### Comment 2: Proportion of twins

The same test gives us the odds of mothers having more children after a twin birth.

```{r}
# Extract coefficients
coef <- coef(logit_fit_EE)

# Calculate odds ratios
odds_ratios <- exp(coef)
```
For the Estonian dataset the  odds of a birth event being the last one for a mother are `r round(odds_ratios[2], 3)` times higher for a twin birth than for a singleton birth.

```{r}
# Extract coefficients
coef <- coef(logit_fit)

# Calculate odds ratios
odds_ratios <- exp(coef)

```
For the other dataset the odds of a birth event being the last one for a mother are `r round(odds_ratios[2], 3)` times higher for a twin birth than for a singleton birth.

## Comment 3: Original Data on Plots

The reviewers also asked to put the data on the Figure 1.

```{r}
# import the function to do model fit and predictions
source("./R/fit_models.R")
```

```{r}
## full data

## Estonia - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_fit <- fitPredictions(dmm_EE, formula, predict = T)

## TwinR - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_fit <- fitPredictions(dmm_orig, formula, predict=T)

```

```{r}
library(ggplot2)
#some nice colors
bc <- c("azure4","purple4", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")

#use new base theme that displays also grid lines
source("./R/twinR_theme.R")
```

### Fig 1a: Estonian vs TwinR Full Data

```{r}

fig2_EE_plot_data <- dmm_EE_fit$results
fig2_orig_plot_data <- dmm_orig_fit$results
```

```{r, fig.height=11.7/2, fig.width=8.3}
infoTxt <- "model prediction with 95% CI and data with values"

fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data,
              aes(y = estimate,  x=births_total, color=estLab),  size = 1) +
  geom_point(data=dmm_EE[dmm_EE$births_total <17, ],
               aes(x=births_total, y=prob_twin, color=estLab, fill = estLab),
               alpha=0.1,
               position = position_nudge(x = -0.1)) +
    stat_summary(data=dmm_EE[dmm_EE$births_total <17, ],
                 aes(x=births_total, y=prob_twin, color=estLab, fill = estLab),
                 alpha=0.1,
                 position = position_jitter(),
                 fun.data=mean_se) +

    geom_ribbon(data=fig2_EE_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    color=estLab, fill = estLab),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data,
             aes(y = estimate,  x=births_total, color=westLab),  size = 1) +
  geom_point(data=dmm_orig[dmm_orig$births_total <19, ],
               aes(x=births_total, y=prob_twin, color=estLab, fill = estLab),
               alpha=0.1,
               position = position_jitter()) +
    stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                 aes(x=births_total, y=prob_twin,
                     color=westLab, fill=westLab),
                 alpha=1,
                 shape = 1,
                 position = position_nudge(x = 0.1),
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill=westLab),
                alpha = 0.1) +
    ggplot2::scale_x_continuous(breaks = 1:18) +

    ggplot2::coord_cartesian() +
    labs(subtitle = paste0("(a) All births, ", infoTxt),
         y="Per-birth twin. prob.",
         x="Maternal total births")
p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "population")

p2
```

The first plot clearly illustrates that we cannot plot the whole data itself on the plot, as there are women who only gave birth to twins. We can plot a subset (eg those with at least 10 children) but this would lose the point of the plot. There is just so much data that is semi-categorical that we cannot think of a way put into a plot so that it gives additional information.

## Comment 4: Dizygotic Twins

## Comment 5: Unmarried Mothers

```{r}
#END
```

### 2023-12-27 S3_response_to_reviewers.Rmd
added precentages

@@
 
+```{r}
+#compare the precentage of twin births between last births and not last births
+cat("Estonian dataset:\n")
+tbl <- table(data_births_monthly_EE[c("twin", "last")])
+tbl
+cat("\n")
+tbl <- tbl / sum(tbl)
+#now calcualte the % by hand
+cat("Last twins: ", round(100 * tbl[2,2] / sum(tbl[,2]), 2), "%\n")
+cat("Prior twins: ",round(100 * tbl[2,1] / sum(tbl[,1]), 2), "%\n")
+
+```
+
+```{r}
+#compare the precentage of twin births between last births and not last births
+cat("Western dataset:\n")
+tbl <- table(data_births_monthly[c("twin", "last")])
+tbl
+cat("\n")
+tbl <- tbl / sum(tbl)
+#now calcualte the % by hand
+cat("Last twins: ", round(100 * tbl[2,2] / sum(tbl[,2]), 2), "%\n")
+cat("Prior twins: ",round(100 * tbl[2,1] / sum(tbl[,1]), 2), "%\n")
+
+```
+
 ## Comment 3: Original Data on Plots

### 2023-12-27 S3_response_to_reviewers.Rmd
fix

--- 2023-12-27 S3_response_to_reviewers.Rmd
@@
 
+```{r}
+#compare the precentage of twin births between last births and not last births
+cat("Estonian dataset:\n")
+tbl <- table(data_births_monthly_EE[c("twin", "last")])
+tbl
+cat("\n")
+#now calcualte the % by hand
+cat("Last twins: ", round(100 * tbl[2,2] / sum(tbl[,2]), 2), "%\n")
+cat("Prior twins: ",round(100 * tbl[2,1] / sum(tbl[,1]), 2), "%\n")
+
+```
+
+```{r}
+#compare the precentage of twin births between last births and not last births
+cat("Western dataset:\n")
+tbl <- table(data_births_monthly[c("twin", "last")])
+tbl
+cat("\n")
+#now calcualte the % by hand
+cat("Last twins: ", round(100 * tbl[2,2] / sum(tbl[,2]), 2), "%\n")
+cat("Prior twins: ",round(100 * tbl[2,1] / sum(tbl[,1]), 2), "%\n")
+
+```
+
 ## Comment 3: Original Data on Plots

### 2024-01-02 R/fit_models.R
added reponse to reviewers

@@ -56,11 +56,12 @@ fitPredictions <- function(dataset, formula,
   get_predictions <- function(predDataFname, fit, dataset, args, save){
     if(!file.exists(predDataFname)){
       nd <- NULL
+      min_births <- min(dataset$births_total)
+      max_births <- max(dataset$births_total)
+      births <- max_births-min_births + 1
       
       # mother level data
       if("births_total" %in% all.vars(args$formula)){
-        min_births <- min(dataset$births_total)
-        max_births <- max(dataset$births_total)
         nd <- data.frame(births_total = min_births:max_births)
       }
       #birth level data
@@ -69,6 +70,15 @@ fitPredictions <- function(dataset, formula,
         ## remove the twin covariate, which is not used in twin_fit:
         nd  <-  nd[nd$twin,] 
         nd$twin <- NULL
+      }
+      
+      if('illigimate_birth' %in% all.vars(args$formula)){
+        nd <- data.frame(births_total = rep(min_births:max_births, 2),
+                         illigimate_birth = c(rep(T, births), rep(F, births)))
+      }
+      if('illigimate_conception' %in% all.vars(args$formula)){
+        nd <- data.frame(births_total = rep(min_births:max_births, 2),
+                         illigimate_conception = c(rep(T, births), rep(F, births)))
       }
       
       if(!is.null(nd)){

### 2024-01-03 S5_response_to_reviewers.Rmd
response 2 reviewers

--- 2023-12-27 S3_response_to_reviewers.Rmd
@@
 ```
+
 ### Comment 2: Proportion of twins
@@
 For the other dataset the odds of a birth event being the last one for a mother are `r round(odds_ratios[2], 3)` times higher for a twin birth than for a singleton birth.
+
+```{r}
+#compare the precentage of twin births between last births and not last births
+cat("Estonian dataset:\n")
+tble <- table(data_births_monthly_EE[c("twin", "last")])
+tble
+cat("\n")
+#now calcualte the % by hand
+cat("Last twins: ", round(100 * tble[2,2] / sum(tble[,2]), 2), "%\n")
+cat("Prior twins: ",round(100 * tble[2,1] / sum(tble[,1]), 2), "%\n")
+
+```
+
+```{r}
+#compare the precentage of twin births between last births and not last births
+cat("Western dataset:\n")
+tblw <- table(data_births_monthly[c("twin", "last")])
+tblw
+cat("\n")
+#now calcualte the % by hand
+cat("Last twins: ", round(100 * tblw[2,2] / sum(tblw[,2]), 2), "%\n")
+cat("Prior twins: ",round(100 * tblw[2,1] / sum(tblw[,1]), 2), "%\n")
+
+```
+
+```{r}
+# Combine the two tables into a 3-dimensional array
+tbl <- array(c(tble, tblw), dim = c(dim(tble), 2))
+
+#numeric problem workaround
+tbl <- tbl / 1
+
+# Perform the Mantel-Haenszel test
+mantelhaen.test(tbl)
+```
 
@@
 The first plot clearly illustrates that we cannot plot the whole data itself on the plot, as there are women who only gave birth to twins.
-We can plot a subset (eg those with at least 10 children) but this would lose the point of the plot.
+We can plot a subset (e.g.
+those with at least 10 children) but this would lose the point of the plot.
 There is just so much data that is semi-categorical that we cannot think of a way put into a plot so that it gives additional information.
@@
 
-## Comment 5: Unmarried Mothers
+We can do the same analysis for dizygotic twins.
+
+```{r}
+#re-import the data
+dbm_EE <- readRDS("./data/data_births_all_EE.rds")
+```
+
+```{r}
+#take out mothers that had twins of single sex
+single_sex_twin_mothers <- as.character(unlist(unique(
+  dbm_EE[dbm_EE$twin & dbm_EE$cf_sex != "mixed", "maternal_id"])
+  ))
+
+dbmEE_wo_single_sex_twins <- dbm_EE[!(dbm_EE$maternal_id %in% 
+                                        single_sex_twin_mothers),]
+```
+
+```{r}
+#the twinR package expects population to be present
+dbmEE_wo_single_sex_twins$pop <- "Estonia"
+
+## Expand the birth level data for the fit of statistical models:
+dbmEE_wo_single_sex_twins <- twinR::expand_data(dbmEE_wo_single_sex_twins)
+
+dbmEE_wo_single_sex_twins <- add_last_birth(dbmEE_wo_single_sex_twins)
+
+dbmEE_wo_single_sex_twins_not_last <- dbmEE_wo_single_sex_twins[!dbmEE_wo_single_sex_twins$last,]
+
+#make the aggregates
+dmm_EE_mixed <- twinR::aggregate_data(dbmEE_wo_single_sex_twins)
+dmm_EE_mixed$prob_twin <- dmm_EE_mixed$twin_total / dmm_EE_mixed$births_total
+
+dmm_EE_mixed_nl <- twinR::aggregate_data(dbmEE_wo_single_sex_twins_not_last)
+dmm_EE_mixed_nl$prob_twin <- dmm_EE_mixed_nl$twin_total / dmm_EE_mixed_nl$births_total
+```
+
+### Full Data
+
+```{r}
+options(scipen = 99)
+```
+
+```{r}
+## Estonia
+formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
+dmm_EE_mixed_fit <- fitPredictions(dmm_EE_mixed, formula)
+
+knitr::kable(build_fit_summary.table(dmm_EE_mixed_fit$fit))
+```
+
+### No Last Births Data
+
+```{r}
+## Estonia
+formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
+dmm_EE_mixed_nl_fit <- fitPredictions(dmm_EE_mixed_nl, formula)
+
+knitr::kable(build_fit_summary.table(dmm_EE_mixed_nl_fit$fit))
+```
+
+### Plot the Predictions
+
+```{r}
+fig2_EE_plot_data <- dmm_EE_mixed_fit$results
+fig2_EE_plot_data_nl <- dmm_EE_mixed_nl_fit$results
+fig2_EE_plot_data_nl$births_total <- fig2_EE_plot_data_nl$births_total + 1
+```
+
+```{r}
+dmm_EE_mixed_nl_plot <- dmm_EE_mixed_nl
+dmm_EE_mixed_nl_plot$births_total <- dmm_EE_mixed_nl_plot$births_total + 1
+```
+
+```{r}
+#colors
+bc <- c("purple", "black")
+fig2_ext_orig <- ggplot() +  
+  geom_line(data=fig2_EE_plot_data,
+              aes(y = estimate,  x=births_total, color="EE all"),
+            linewidth = 1) +
+    stat_summary(data=dmm_EE_mixed[dmm_EE_mixed$births_total <14, ],
+                 aes(x=births_total, y=prob_twin, color="EE all",
+                     fill = "EE all"), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=fig2_EE_plot_data,
+                aes(y = estimate,  x=births_total,
+                    ymin = lwr, ymax = upr, color="EE all", fill = "EE all"),
+                alpha = 0.3) + 
+    geom_line(data=fig2_EE_plot_data_nl,
+              aes(y = estimate,  x=births_total, color="EE w/o last"),
+              linewidth = 1) +
+    stat_summary(data=dmm_EE_mixed_nl_plot[dmm_EE_mixed_nl_plot$births_total <14,],
+                 aes(x=births_total, y=prob_twin, color="EE w/o last",
+                     fill = "EE w/o last"), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=fig2_EE_plot_data_nl,
+                aes(y = estimate,  x=births_total, ymin = lwr,
+                    ymax = upr, color="EE w/o last", fill = "EE w/o last"),
+                alpha = 0.3) + 
+    ggplot2::scale_x_continuous(breaks = 1:18) + 
+    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.002)) +
+    ggplot2::coord_cartesian() +
+    labs(subtitle = "Prediction only mixed sex twins + data mean with SE",
+         y="Per-birth twin. prob.",
+         x="Maternal total births")
+p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")
+
+p2 
+```
+
+The mixed sex twinning rate seem to remain constant if last births are excluded from the data and slightly decline if last births are included. The decline seems not to be significant.
+
+## Comment 5: Illegitimate Maternities
+
+We can  add a predictor to the model describing if the women had first birth or first conception before being married. We will do both to investigate if illegitimate maternities affect twinning rate.
+
+```{r}
+#add the illigimate children columns
+illigimate_birth <- unique(data_births_monthly_EE[c( "illigimate_birth",
+                                                     "maternal_id")])
+illigimate_birth <- setNames(illigimate_birth$illigimate_birth,
+                             illigimate_birth$maternal_id)
+
+illigimate_conception <- unique(data_births_monthly_EE[c( "illigimate_conception","maternal_id")])
+illigimate_conception <- setNames(illigimate_conception$illigimate_conception, illigimate_conception$maternal_id)
+```
+
+```{r}
+dmm_EE$illigimate_conception <- illigimate_conception[dmm_EE$maternal_id]
+
+dmm_EE$illigimate_birth <- illigimate_birth[dmm_EE$maternal_id]
+```
+
+```{r}
+dmm_EE_nl$illigimate_conception <- illigimate_conception[dmm_EE_nl$maternal_id]
+
+dmm_EE_nl$illigimate_birth <- illigimate_birth[dmm_EE_nl$maternal_id]
+```
+
+```{r}
+cat("First birth before marriage:\n")
+table(dmm_EE$illigimate_birth)
+```
+
+```{r}
+cat("First conception before marriage:\n")
+table(dmm_EE$illigimate_conception)
+```
+
+### With All Data
+
+```{r}
+## Estonia
+formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + illigimate_birth"
+dmm_EE_fit_birth <- fitPredictions(dmm_EE, formula)
+
+knitr::kable(build_fit_summary.table(dmm_EE_fit_birth$fit))
+```
+
+```{r}
+## Estonia
+formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + illigimate_conception"
+dmm_EE_fit_conception <- fitPredictions(dmm_EE, formula)
+
+knitr::kable(build_fit_summary.table(dmm_EE_fit_conception$fit))
+```
+
+### No Last Births Data
+
+```{r}
+## Estonia
+formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + illigimate_birth"
+dmm_EE_nl_fit_birth <- fitPredictions(dmm_EE_nl, formula)
+
+knitr::kable(build_fit_summary.table(dmm_EE_nl_fit_birth$fit))
+```
+
+```{r}
+## Estonia
+formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + illigimate_conception"
+dmm_EE_fit_nl_conception <- fitPredictions(dmm_EE_nl, formula)
+
+knitr::kable(build_fit_summary.table(dmm_EE_fit_nl_conception$fit))
+```
+
+### Figures for Estonian mothers with and without Last Birth
+
+#### Illigimate Birth
+
+```{r}
+fig2_EE_plot_data <- dmm_EE_fit_birth$results
+fig2_EE_plot_data_nl <- dmm_EE_nl_fit_birth$results
+fig2_EE_plot_data_nl$births_total <- fig2_EE_plot_data_nl$births_total + 1
+```
+
+```{r}
+dmm_EE_nl_plot <- dmm_EE_nl
+dmm_EE_nl_plot$births_total <- dmm_EE_nl_plot$births_total + 1
+```
+
+```{r, fig.width=6, fig.height=4}
+#colors
+bc <- c("purple", "black")
+fig2_ext_orig <- ggplot() +  
+  geom_line(data=fig2_EE_plot_data,
+              aes(y = estimate,  x=births_total, color=illigimate_birth),
+            linewidth = 1) +
+    stat_summary(data=dmm_EE[dmm_EE$births_total <13, ],
+                 aes(x=births_total, y=prob_twin, color=illigimate_birth,
+                     fill = illigimate_birth), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=fig2_EE_plot_data,
+                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
+                    color= illigimate_birth, fill = illigimate_birth),
+                alpha = 0.3) +
+    ggplot2::scale_x_continuous(breaks = 1:18) + 
+    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
+    ggplot2::coord_cartesian() +
+    labs(subtitle = "Prediction from full data model + data mean with SE",
+         y="Per-birth twin. prob.",
+         x="Maternal total births")
+p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+  scale_fill_manual(values=bc) + guides(color="none") +
+  labs(fill = "Illigimate I birth")
+cat("\n")
+p2 
+cat("\n")
+```
+
+```{r}
+#colors
+bc <- c("purple", "black")
+fig2_ext_orig <- ggplot() +  
+  geom_line(data=fig2_EE_plot_data_nl,
+              aes(y = estimate,  x=births_total, color=illigimate_birth), 
+            linewidth = 1) +
+    stat_summary(data=dmm_EE_nl[dmm_EE_nl$births_total <12, ],
+                 aes(x=births_total, y=prob_twin, color=illigimate_birth,
+                     fill = illigimate_birth), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=fig2_EE_plot_data_nl,
+                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
+                    color= illigimate_birth, fill = illigimate_birth),
+                alpha = 0.3) +
+    ggplot2::scale_x_continuous(breaks = 1:18) + 
+    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
+    ggplot2::coord_cartesian() +
+    labs(subtitle = "Prediction from w/o last model + data mean with SE",
+         y="Per-birth twin. prob.",
+         x="Maternal total births")
+p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+  scale_fill_manual(values=bc) + guides(color="none") +
+  labs(fill = "Illigimate I birth")
+cat("\n")
+p2 
+cat("\n")
+```
+
+#### Illigimate conception
+
+```{r}
+fig2_EE_plot_data <- dmm_EE_fit_conception$results
+fig2_EE_plot_data_nl <- dmm_EE_fit_nl_conception$results
+fig2_EE_plot_data_nl$births_total <- fig2_EE_plot_data_nl$births_total + 1
+```
+
+```{r}
+dmm_EE_nl_plot <- dmm_EE_nl
+dmm_EE_nl_plot$births_total <- dmm_EE_nl_plot$births_total + 1
+```
+
+```{r, fig.width=6, fig.height=4}
+#colors
+bc <- c("purple", "black")
+fig2_ext_orig <- ggplot() +  
+  geom_line(data=fig2_EE_plot_data,
+              aes(y = estimate,  x=births_total, color=illigimate_conception),
+            linewidth = 1) +
+    stat_summary(data=dmm_EE[dmm_EE$births_total <13, ],
+                 aes(x=births_total, y=prob_twin, color=illigimate_conception,
+                     fill = illigimate_conception), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=fig2_EE_plot_data,
+                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
+                    color= illigimate_conception, fill = illigimate_conception),
+                alpha = 0.3) +
+    ggplot2::scale_x_continuous(breaks = 1:18) + 
+    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
+    ggplot2::coord_cartesian() +
+    labs(subtitle = "Prediction from full data model + data mean with SE",
+         y="Per-birth twin. prob.",
+         x="Maternal total births")
+p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+  scale_fill_manual(values=bc) + guides(color="none") +
+  labs(fill = "Illigimate I conception")
+cat("\n")
+p2 
+cat("\n")
+```
+
+```{r}
+#colors
+bc <- c("purple", "black")
+fig2_ext_orig <- ggplot() +  
+  geom_line(data=fig2_EE_plot_data_nl,
+              aes(y = estimate,  x=births_total, color=illigimate_conception),
+            linewidth = 1) +
+    stat_summary(data=dmm_EE_nl[dmm_EE_nl$births_total <12, ],
+                 aes(x=births_total, y=prob_twin, color=illigimate_conception,
+                     fill = illigimate_conception), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=fig2_EE_plot_data_nl,
+                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
+                    color= illigimate_conception, fill = illigimate_conception),
+                alpha = 0.3) +
+    ggplot2::scale_x_continuous(breaks = 1:18) + 
+    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
+    ggplot2::coord_cartesian() +
+    labs(subtitle = "Prediction from w/o last model + data mean with SE",
+         y="Per-birth twin. prob.",
+         x="Maternal total births")
+p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+  scale_fill_manual(values=bc) + guides(color="none") + 
+  labs(fill = "Illigimate I conception")
+cat("\n")
+p2
+cat("\n")
+```
 

### 2024-01-03 S5_response_to_reviewers.Rmd
pretify

@@
 
-### Figures for Estonian mothers with and without Last Birth
+### Illigimate Birth Figures for Estonian mothers With and Without Last Birth
 
-#### Illigimate Birth
+```{r}
+#' Plot Binary Predictions
+#'
+#' This function generates a ggplot based on the provided parameters.
+#'
+#' @param plot_data The data to be plotted.
+#' @param summary_data The data to be summarized.
+#' @param color_var The color variable for the plot.
+#' @param threshold The threshold for births_total mean data to display.
+#'
+#' @return A ggplot object.
+#'
+plot_predictions <- function(plot_data, summary_data, color_var, threshold = 13) {
+  fig2_ext_orig <- ggplot() +  
+    geom_line(data=plot_data,
+              aes_string(y = "estimate",  x="births_total", color=color_var),
+              linewidth = 1) +
+    stat_summary(data=summary_data[summary_data$births_total < threshold, ],
+                 aes_string(x="births_total", y="prob_twin", color=color_var,
+                            fill = color_var), alpha=0.5,
+                 fun.data=mean_se) +
+    geom_ribbon(data=plot_data,
+                aes_string(y = "estimate",  x="births_total", ymin = "lwr",
+                           ymax = "upr",
+                           color=color_var, fill = color_var),
+                alpha = 0.3) +
+    ggplot2::scale_x_continuous(breaks = 1:18) + 
+    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
+    ggplot2::coord_cartesian()
+  return(fig2_ext_orig)
+}
+
+```
 
@@
 
-```{r, fig.width=6, fig.height=4}
-#colors
-bc <- c("purple", "black")
-fig2_ext_orig <- ggplot() +  
-  geom_line(data=fig2_EE_plot_data,
-              aes(y = estimate,  x=births_total, color=illigimate_birth),
-            linewidth = 1) +
-    stat_summary(data=dmm_EE[dmm_EE$births_total <13, ],
-                 aes(x=births_total, y=prob_twin, color=illigimate_birth,
-                     fill = illigimate_birth), alpha=0.5,
-                 fun.data=mean_se) +
-    geom_ribbon(data=fig2_EE_plot_data,
-                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color= illigimate_birth, fill = illigimate_birth),
-                alpha = 0.3) +
-    ggplot2::scale_x_continuous(breaks = 1:18) + 
-    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
-    ggplot2::coord_cartesian() +
+```{r}
+plot_predictions(plot_data = fig2_EE_plot_data,
+                 summary_data = dmm_EE,
+                 color_var = "illigimate_birth") +
   labs(subtitle = "Prediction from full data model + data mean with SE",
          y="Per-birth twin. prob.",
-         x="Maternal total births")
-p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+         x="Maternal total births") +
+  base_theme(larger=8) + scale_color_manual(values=bc) + 
   scale_fill_manual(values=bc) + guides(color="none") +
   labs(fill = "Illigimate I birth")
-cat("\n")
-p2 
-cat("\n")
 ```
@@
 ```{r}
-#colors
-bc <- c("purple", "black")
-fig2_ext_orig <- ggplot() +  
-  geom_line(data=fig2_EE_plot_data_nl,
-              aes(y = estimate,  x=births_total, color=illigimate_birth), 
-            linewidth = 1) +
-    stat_summary(data=dmm_EE_nl[dmm_EE_nl$births_total <12, ],
-                 aes(x=births_total, y=prob_twin, color=illigimate_birth,
-                     fill = illigimate_birth), alpha=0.5,
-                 fun.data=mean_se) +
-    geom_ribbon(data=fig2_EE_plot_data_nl,
-                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color= illigimate_birth, fill = illigimate_birth),
-                alpha = 0.3) +
-    ggplot2::scale_x_continuous(breaks = 1:18) + 
-    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
-    ggplot2::coord_cartesian() +
+plot_predictions(plot_data = fig2_EE_plot_data_nl,
+                 summary_data = dmm_EE_nl_plot,
+                 color_var = "illigimate_birth") +
     labs(subtitle = "Prediction from w/o last model + data mean with SE",
          y="Per-birth twin. prob.",
-         x="Maternal total births")
-p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+         x="Maternal total births")+ 
+  base_theme(larger=8) + scale_color_manual(values=bc) + 
   scale_fill_manual(values=bc) + guides(color="none") +
   labs(fill = "Illigimate I birth")
-cat("\n")
-p2 
-cat("\n")
+
 ```
 
-#### Illigimate conception
+### Illigimate Conception Figures for Estonian mothers With and Without Last Birth
 
@@
 ```{r}
-dmm_EE_nl_plot <- dmm_EE_nl
-dmm_EE_nl_plot$births_total <- dmm_EE_nl_plot$births_total + 1
-```
-
-```{r, fig.width=6, fig.height=4}
-#colors
-bc <- c("purple", "black")
-fig2_ext_orig <- ggplot() +  
-  geom_line(data=fig2_EE_plot_data,
-              aes(y = estimate,  x=births_total, color=illigimate_conception),
-            linewidth = 1) +
-    stat_summary(data=dmm_EE[dmm_EE$births_total <13, ],
-                 aes(x=births_total, y=prob_twin, color=illigimate_conception,
-                     fill = illigimate_conception), alpha=0.5,
-                 fun.data=mean_se) +
-    geom_ribbon(data=fig2_EE_plot_data,
-                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color= illigimate_conception, fill = illigimate_conception),
-                alpha = 0.3) +
-    ggplot2::scale_x_continuous(breaks = 1:18) + 
-    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
-    ggplot2::coord_cartesian() +
+plot_predictions(plot_data = fig2_EE_plot_data,
+                 summary_data = dmm_EE,
+                 color_var = "illigimate_conception",
+                 threshold = 12) +
   labs(subtitle = "Prediction from full data model + data mean with SE",
          y="Per-birth twin. prob.",
-         x="Maternal total births")
-p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+         x="Maternal total births") +
+  base_theme(larger=8) + scale_color_manual(values=bc) + 
   scale_fill_manual(values=bc) + guides(color="none") +
   labs(fill = "Illigimate I conception")
-cat("\n")
-p2 
-cat("\n")
 ```
@@
 ```{r}
-#colors
-bc <- c("purple", "black")
-fig2_ext_orig <- ggplot() +  
-  geom_line(data=fig2_EE_plot_data_nl,
-              aes(y = estimate,  x=births_total, color=illigimate_conception),
-            linewidth = 1) +
-    stat_summary(data=dmm_EE_nl[dmm_EE_nl$births_total <12, ],
-                 aes(x=births_total, y=prob_twin, color=illigimate_conception,
-                     fill = illigimate_conception), alpha=0.5,
-                 fun.data=mean_se) +
-    geom_ribbon(data=fig2_EE_plot_data_nl,
-                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
-                    color= illigimate_conception, fill = illigimate_conception),
-                alpha = 0.3) +
-    ggplot2::scale_x_continuous(breaks = 1:18) + 
-    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
-    ggplot2::coord_cartesian() +
+plot_predictions(plot_data = fig2_EE_plot_data_nl,
+                 summary_data = dmm_EE_nl_plot,
+                 color_var = "illigimate_conception",
+                 threshold = 12) +
     labs(subtitle = "Prediction from w/o last model + data mean with SE",
          y="Per-birth twin. prob.",
-         x="Maternal total births")
-p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
+         x="Maternal total births")+ 
+  base_theme(larger=8) + scale_color_manual(values=bc) + 
   scale_fill_manual(values=bc) + guides(color="none") +
   labs(fill = "Illigimate I conception")
-cat("\n")
-p2
-cat("\n")
+
 ```

### 2024-01-04 S1_replicate_stats_and_plots.Rmd
rename title

@@
 ---
-title: 'Suplementary Material 1: Code to Reproduce Analysis and Plots'
+title: 'Suplementary Code 1: Code to Reproduce Analysis and Plots'
 author: "Richard Meitern"

### 2024-01-04 S2_draw_publication_plots.Rmd
rename title

@@
 ---
-title: 'Suplementary Material 2: Drawing Publication Plots'
+title: 'Suplementary Code 2: Drawing Publication Plots'
 author: "Richard Meitern"

### 2024-01-04 S3_run_simulations_EE_not_last.Rmd
rename title

@@
 ---
-title: 'Suplementary Material 3: Run Simulations on the Last Birth Omited Data'
+title: 'Suplementary Code 3: Run Simulations on the Last Birth Omited Data'
 author: "Richard Meitern"

### 2024-01-04 S5_response_to_reviewers.Rmd
added odds ratio

@@
 ---
-title: 'Suplementary Material 3: Code to Respond to Reviewers'
+title: 'Suplementary Code 5: Code to Respond to Reviewers'
 author: "Richard Meitern"
@@
 #fit logistic regression Estonia
-logit_fit_EE <- glm(last ~ twin,
+logit_fit_EE <- glm(twin ~  last,
                  data = data_births_monthly_EE,
                  family = binomial(link = "logit"))
-summary(logit_fit_EE)
+summary_logit_fit_EE <-  summary(logit_fit_EE)
+summary_logit_fit_EE
 ```
@@
 #fit logistic regression Other European
-logit_fit <- glm(last ~ twin,
+logit_fit <- glm(twin ~  last,
                  data = data_births_monthly,
                  family = binomial(link = "logit"))
-summary(logit_fit)
+summary_logit_fit_W <-  summary(logit_fit)
+summary_logit_fit_W
 ```
@@
 
+#### Test the Interaction
+
 ```{r}
-# Combine the two tables into a 3-dimensional array
-tbl <- array(c(tble, tblw), dim = c(dim(tble), 2))
+data_births_monthly$monthly <- NULL
+#data_births_monthly$pop <- "Western"
+dbmFull <- rbind(data_births_monthly_EE[colnames(data_births_monthly)],
+                 data_births_monthly)
+dbmFull$popEE <- dbmFull$pop == "Estonia"
+```
 
-#numeric problem workaround
-tbl <- tbl / 1
+```{r}
+#fit logistic regression
+logit_fit_full <- glm(twin ~  last + popEE + popEE:last,
+                 data = dbmFull,
+                 family = binomial(link = "logit"))
+summary_logit_fit_full <-  summary(logit_fit_full)
+summary_logit_fit_full
 
-# Perform the Mantel-Haenszel test
-mantelhaen.test(tbl)
+```
+
+```{r}
+#' Get Coefficient Table
+#'
+#' This function returns a table with odds ratios and their confidence intervals.
+#'
+#' @param summary A summary object from a glm model.
+#' @param datasetname A character string specifying the name of the dataset.
+#'
+#' @return A data frame with odds ratios and their confidence intervals.
+get_coef_tbl <- function(summary, datasetname){
+  tbl <- as.data.frame(summary$coefficients)
+  tbl$dataset <- datasetname
+  tbl$variable <- rownames(tbl)
+  tbl$`Odds Ratio` <- exp(tbl$Estimate)
+  tbl$`CI low` <- exp(tbl$Estimate - 1.96 * tbl$`Std. Error`)
+  tbl$`CI high` <- exp(tbl$Estimate + 1.96 * tbl$`Std. Error`)
+  rownames(tbl) <- NULL
+  tbl <- tbl[,c(5,6,7:9)]
+  return(tbl)
+}
+
+#print all coeficient tables in one table
+#add dataset name to each table
+coefEEtbl <- get_coef_tbl(summary_logit_fit_EE, "Estonia")
+ceofWtbl <- get_coef_tbl(summary_logit_fit_W, "Western")
+ceofFulltbl <- get_coef_tbl(summary_logit_fit_full, "Combined")
+
+tbl <- rbind(coefEEtbl, ceofWtbl, ceofFulltbl)
+knitr::kable(tbl, digits = 3, caption = "Odds ratios and confidence intervals for the logistic regression model with the response variable being the binary variable *twin* and the predictor variable being the binary variable *last*.")
+```
+```{r}
+#print the frequency table
+frqTbl <- aggregate(maternal_id ~ twin + last + popEE,
+                    data = dbmFull, FUN = length)
+
+# Rename the variables inside the columns
+frqTbl$twin <- ifelse(frqTbl$twin == TRUE, "twins", "singleton")
+frqTbl$last <- ifelse(frqTbl$last == TRUE, "Yes", "No")
+frqTbl$popEE <- ifelse(frqTbl$popEE == TRUE, "Estonia", "Western")
+
+# Rename the column names
+colnames(frqTbl) <- c("Birth", "Last Breeding", "Population", "Count")
+knitr::kable(frqTbl, digits = 0, caption = "Frequency table of the data used for the logistic regression model with the response variable being the binary variable *twin* and the predictor variable being the binary variable *last*.")
+
 ```

### 2024-01-04 S4_draw_IBI_predictions_EE_not_last.Rmd

--- 2023-04-18 S4_draw_IBI_predictions_EE_not_last.Rmd
@@
 ---
-title: 'Suplementary Material 4: Draw Interbirth Interval Predictons on the Last Birth Omited Data'
+title: 'Suplementary Code 4: Draw Interbirth Interval Predictons on the Last Birth Omited Data'
 author: "Richard Meitern"

### 2024-01-09 S5_response_to_reviewers.Rmd
fixed text

@@
 
-## Fitting models
-
-### Comment 1: Logistic regression
+## Comment 1: Logistic regression
 
@@
 #fit logistic regression Estonia
-logit_fit_EE <- glm(twin ~  last,
+logit_fit_EE <- glm(last ~  twin,
                  data = data_births_monthly_EE,
@@
 #fit logistic regression Other European
-logit_fit <- glm(twin ~  last,
+logit_fit <- glm(last  ~  twin,
                  data = data_births_monthly,
@@
 
-### Comment 2: Proportion of twins
+## Comment 2: Proportion of twins
 
@@
 
-#### Test the Interaction
+#### Test the Population Interaction
+
+We check if the population has an effect in a reversed model where the predictor variable is the binary variable *last* and the response variable is the binary variable *twin*.
+
+```{r}
+#fit logistic regression Estonia
+logit_fit_EE <- glm(twin ~  last,
+                 data = data_births_monthly_EE,
+                 family = binomial(link = "logit"))
+summary_logit_fit_EE <-  summary(logit_fit_EE)
+summary_logit_fit_EE
+```
+
+```{r}
+#fit logistic regression Other European
+logit_fit <- glm(twin ~  last,
+                 data = data_births_monthly,
+                 family = binomial(link = "logit"))
+summary_logit_fit_W <-  summary(logit_fit)
+summary_logit_fit_W
+```
 
@@
 
-The first plot clearly illustrates that we cannot plot the whole data itself on the plot, as there are women who only gave birth to twins.
-We can plot a subset (e.g.
-those with at least 10 children) but this would lose the point of the plot.
-There is just so much data that is semi-categorical that we cannot think of a way put into a plot so that it gives additional information.
+The plot illustrates that we cannot plot the raw data along with model predictions, as such figure would be un-informative and highly confusing.
+We cannot think of a way about how to put such an amount of semi-categorical data into a plot so that it would give additional information.
 
-## Comment 4: Dizygotic Twins
+## Comment 4: Opposite-sex Twins
 
-We can do the same analysis for dizygotic twins.
+We can do the same analysis for opposite-sex  twins.
 
@@
 
-The mixed sex twinning rate seem to remain constant if last births are excluded from the data and slightly decline if last births are included. The decline seems not to be significant.
+The mixed sex twinning rate is independent on the number of total births if last births are excluded from the data.
+If last births are included, the association between mixed sex twinning rate and number of total births is negative.
 
